---
title: "Isauro R"
format: 
  html: 
    toc: true
    toc-depth: 3
    embed-resources: true
    code-overflow: wrap
    page-layout: full
---

```{r}
# load ggplot2
library(ggplot2)
library(here)
library(tidyverse)
library(plotly)

source(here::here("data", "data_cleaning.R"))
```

## EDA 1: Average Number of Firms

### By Types of Tech

```{r}
tech_2018_clean |>
  filter(!is.na(n_firms)) |>
  group_by(tech_type) |>
  summarize(avgFirms = mean(n_firms)) |>
  arrange(desc(avgFirms))
```

> Among Tech Companies in the North America, `Specialized Software	` companies have the highest average number of firms at `1324.265`. `Specialized Equipment` companies report the lowest at `1189.696	` firms.

### By Tech Usage

```{r}
tech_2018_clean |>
  filter(!is.na(n_firms)) |>
  group_by(used_tech) |>
  summarize(avgFirms = mean(n_firms)) |>
  arrange(desc(avgFirms))
```

> Among Tech Companies in the North America, Companies that `Did not use` their tech have the highest average number of firms at `4514.4706`. Companies that `Tested` their tech report the lowest at `144.5495` firms.

### By Types of Tech and If They Were Used

```{r}
tech_2018_clean |>
  filter(!is.na(n_firms)) |>
  group_by(tech_type, used_tech) |>
  summarize(avgFirms = mean(n_firms)) |>
  arrange(desc(avgFirms))
```

> Among Tech Companies in the North America, `Robotics` companies that `Did not use` their tech have the highest average number of firms at `5360.76368`. `Robotics` companies that `Tested` their tech report the lowest at `87.01887` firms.

## EDA 2: Average Payroll per Employee

### North American Industry Classification System (naics_sector)

**Note:** `naics_sector` indices and their respective industry:

-   Total = `00`
-   Agriculture, Forestry, Fishing and Hunting = `11`
-   Mining, Quarrying, and Oil and Gas Extraction = `21`
-   Utilities = `22`
-   Construction = `23`
-   Manufacturing = `31-33`
-   Wholesale Trade = `42`
-   Retail Trade = `44-45`
-   Transportation and Warehousing = `48-49`
-   Information = `51`
-   Finance and Insurance = `52`
-   Real Estate Rental and Leasing = `53`
-   Professional, Scientific, and Technical Services = `54`
-   Management of Companies and Enterprises = `55`
-   Administrative and Support and Waste Management and Remediation Services = `56`
-   Educational Services = `61`
-   Health Care and Social Assistance = `62`
-   Arts, Entertainment, and Recreation = `71`
-   Accommodation and Food Services = `72`
-   Other Services (except Public Administration) = `81`
-   Public Administration = `92`
-   NA = `99`

> **Source:** https://www.census.gov/naics/?58967?yearbck=2017

```{r}
cs_panel_clean |>
  mutate(payroll_per_emp = payroll/n_emp) |>
  group_by(naics_sector) |>
  summarize(medianPay = median(payroll_per_emp)) |>
  arrange(desc(medianPay))
```

> Among industries in North America, `Utilities` report the highest median payroll per employee at `$105.31`. `Accomodation and Food Services` report the lowest at `$18.67`.

### Year

```{r}
cs_panel_clean |>
  mutate(payroll_per_emp = payroll/n_emp) |>
  group_by(year) |>
  summarize(medianPay = median(payroll_per_emp)) |>
  arrange(desc(medianPay))
```

> `2021` reported the highest median payroll per employee at `$58.71`. `2017` reported the lowest at `$47.78`.

### naics_sector and Year

```{r}
cs_panel_clean |>
  mutate(payroll_per_emp = payroll/n_emp) |>
  group_by(naics_sector, year) |>
  summarize(medianPay = median(payroll_per_emp)) |>
  arrange(desc(medianPay))
```

> Among industries in North America throughout a 5-year period, `Utilities` report the highest median payroll per employee at `$111.38` during `2021`. `Accomodation and Food Services` report the lowest at `$16.18` during `2020`.
