---
title: "David's Exploration"
format: 
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Motivations

In this document, I will be looking at the evolution of AI. Since we only have data for 2018 on AI-usage, I would look within those industries leading the AI race, how much have their payroll evolve. We excluded the *total* observations, as well as the *unknown* category.

## Introduction

```{r, setup, message=FALSE}
library(here)
library(tidyverse)
library(tigris)
library(stringr)
library(scales)
library(ggrepel)
library(tidymodels)
library(ranger)
library(patchwork)
library(leaflet) 
library(forcats)
library(rpart)
library(tibble)   
library(rpart.plot)


source(here::here("data", "data_cleaning.R"))

## Theme Settings
theme_set(
  theme_minimal(base_size = 12) + 
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30"),
      plot.caption = element_text(hjust = 0, color = "grey40", size = 10),
      panel.grid.minor = element_blank()))
```


## Who's the top industry using AI?

```{r, top_ai}
ai_industry_leaders <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence", 
    !is.na(used_tech), 
    used_tech != "Unknown") |>
  group_by(naics_label) |>
  summarise(
    total_firms = sum(n_firms, na.rm = TRUE),
    ai_firms = sum(n_firms[used_tech == "Used"], na.rm = TRUE),
    .groups = "drop") |>
  mutate(ai_rate = ai_firms / total_firms) |>
  filter(total_firms > 0) |>
  arrange(desc(ai_rate))

ai_industry_leaders |> 
  filter(ai_firms>0) |> 
  ggplot(aes(x = reorder(naics_label, ai_rate), y = ai_rate)) +
  geom_col() + 
  scale_fill_viridis_b()+
  coord_flip() + 
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Which Industries Led the AI Race in 2018?",
    subtitle = "Percentage of firms reporting AI Use",
    x = NULL,
    y = "AI Adoption Rate",
    caption = "Management took the lead in 2018, with an AI implementation of 7.7%. \nSource: 2018 Annual Business Survey (ABS) | Author: David Rios ") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(), 
        axis.ticks.y = element_blank())
```
## Payroll Level Change 

```{r}
tech_2018_summary <- tech_2018_clean |> 
  group_by(year, used_tech, use_level) |> 
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop"
  )

payroll_all <- cs_panel_clean |>
  group_by(year, naics_label) |>
  summarise(total_payroll = sum(payroll * 1000, na.rm = TRUE), .groups = "drop")

top_4_industries <- ai_industry_leaders |> slice_max(ai_rate, n = 4) |> pull(naics_label)
payroll_top4 <- payroll_all |> filter(naics_label %in% top_4_industries)

ggplot() +
  geom_line(data = payroll_all, 
            aes(x = year, y = total_payroll, group = naics_label), 
            color = "grey85", linewidth = 0.5, alpha = 0.6) +
  geom_line(data = payroll_top4, 
            aes(x = year, y = total_payroll, color = naics_label, group = naics_label), 
            linewidth = 1) +
  geom_point(data = payroll_top4, 
             aes(x = year, y = total_payroll, color = naics_label), 
             size = 1.5) +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-9, suffix = " B", accuracy = 1)) +
  scale_color_viridis_d(option = "H") +
  labs(
    title = "Payroll Evolution",
    subtitle = "Highlighting the 4 industries with the highest AI adoption rates in 2018",
    x = "Year",
    y = "Total Annual Payroll (Billions)",
    color = NULL,
    caption = "Healthcare is the top industry by total payroll, despite showing low AI adoption rates in 2018.  \nSource: 2018 Annual Business Survey (ABS) | Author: David Rios "
  ) + theme(legend.position = "top")
```

## Payroll Percentage Change

```{r}
## indexed payroll growth
payroll_growth <- payroll_all |>
  arrange(naics_label, year) |>
  group_by(naics_label) |>
  mutate(
    payroll_growth_pct = total_payroll / first(total_payroll) #base year: 2017
  ) |>
  ungroup() |>
  mutate(naics_label = str_wrap(naics_label, width = 35))


## Top AI industries
top_4_industries <- ai_industry_leaders |>
  slice_max(ai_rate, n = 4) |>
  pull(naics_label)

payroll_growth_top4 <- payroll_growth |>
  filter(naics_label %in% top_4_industries)


## Bottom AI industries
w_4_industries <- ai_industry_leaders |>
  slice_min(ai_rate, n = 4) |>
  pull(naics_label)

payroll_growth_w4 <- payroll_growth |>
  filter(naics_label %in% w_4_industries)


# plots
p1_top4_ai <- ggplot() +
  geom_line(
    data = payroll_growth,
    aes(x = year, y = payroll_growth_pct, group = naics_label),
    color = "grey85", linewidth = 0.5, alpha = 0.6) +
  geom_line(
    data = payroll_growth_top4,
    aes(x = year, y = payroll_growth_pct, color = naics_label),
    linewidth = 1) +
  geom_text(
    data = payroll_growth_top4 |> filter(year == max(year)),
    aes(x = year, y = payroll_growth_pct, label = naics_label, color = naics_label),
    nudge_x = 1,
    show.legend = FALSE,
    size = 3) +
  scale_y_continuous(labels = percent_format()) +
    scale_color_viridis_d(option = "D") +
  labs(
    title = "AI-Leading Industries",
    subtitle = "Indexed to 2017 total payroll",
    x = "Year",
    y = "Payroll (2017 = 1)"
  ) +
  theme_minimal() +
 theme(
    legend.position = "none"
  )

p2_worse_ai <- ggplot() +
  geom_line(
    data = payroll_growth,
    aes(x = year, y = payroll_growth_pct, group = naics_label),
    color = "grey85", linewidth = 0.5, alpha = 0.6
  ) +
  geom_line(
    data = payroll_growth_w4,
    aes(x = year, y = payroll_growth_pct, color = naics_label),
    linewidth = 1) +
  geom_point(
    data = payroll_growth_w4,
    aes(x = year, y = payroll_growth_pct, color = naics_label),
    size = 1) +
  geom_text(
    data = payroll_growth_w4 |> filter(year == max(year)),
    aes(
      x = year,
      y = payroll_growth_pct,
      label = naics_label,
      color = naics_label),
    nudge_x = 1,
    show.legend = FALSE,
    size = 3) +
  scale_y_continuous(labels = percent_format()) +
  scale_color_viridis_d(option = "H") +
  labs(
    title = "Lower AI-Adoption Industries",
    subtitle = "Indexed to 2017 total payroll",
    x = "Year",
    y = NULL,
    caption = "Source: 2018 ABS + CS Panel Data"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )

## final plot combining both plots above 
final_dis <- p1_top4_ai + p2_worse_ai +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Payroll Growth Comparison",
    theme = theme(
      plot.title = element_text(face = "bold", size = 14)
    )
  )

final_dis
```


## Revenue

```{r}
revenue_2017 <- cs_panel_clean |>
  filter(year == 2017) |>
  group_by(naics_label) |>
  summarise(total_revenue = sum(revenue * 1000, na.rm = TRUE), .groups = "drop") |>
  filter(total_revenue > 0) |>
  arrange(desc(total_revenue))

#  2017 snapshot (in Trillions)
ggplot(revenue_2017, aes(x = total_revenue, y = reorder(naics_label, total_revenue))) +
  geom_col(fill = "grey30", width = 0.9) + 
  scale_x_continuous(
    labels = scales::label_dollar(scale = 1e-12, suffix = "T", accuracy = 1),
    breaks = seq(0, 10e12, by = 1e12),
    expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Total Industry Revenue in 2017",
    subtitle = "A snapshot of financial volume right before the 2018 AI Survey",
    x = "Total Revenue (Trillions)",
    y = NULL,
    caption = "Source: 2018 Annual Business Survey (ABS) | Author: David Rios ") +  theme_minimal(base_size = 12) + 
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    plot.caption = element_text(hjust = 0, color = "grey40", size = 8),
    panel.grid.minor = element_blank())
```


## The AI Wage Premium

Does the AI wage premium holds constant as firms grow in size?

```{r, wage_premium}
## wage premium story
tech_2018_clean |>
  filter(tech_type == "artificial intelligence",
         used_tech != "Unknown",
         naics_sector != "Total", 
         payroll > 0,
         n_emp > 0) |>
  mutate(
    ai_used = if_else(used_tech == "Used", "AI Adopters", "Non-Adopters"),
    payroll_per_emp = payroll / n_emp) |> 
  ggplot(aes(x = n_emp, y = payroll_per_emp, color = ai_used)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(se = FALSE, linewidth = 1.5) +  # method = "mgcv",
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(
    labels = scales::label_dollar(suffix = "K"), 
    breaks = seq(0, 300, by = 50)) +  
    scale_color_viridis_d(option = "D") +
  labs(
    title = "Does AI Pay?",
    subtitle = "Evidence of higher payroll standards among technology adopters across firm scales",
    x = "Number of Employees (Log Scale)",
    y = "Avg. Payroll per Employee (USD)",
    color = "Adoption Status",
    caption = "Payroll standardized to thousands of dollars per employee. Trend lines calculated via local regression (LOESS). \n Source: 2018 Annual Business Survey (ABS) | Author: David Rios"
  ) +
  theme_minimal() + 
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8))
```


## AI Hubs in the US

```{r}
# AI Adoption Rate by State
state_ai_adoption <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    !is.na(used_tech),
    used_tech != "Unknown"
  ) |>
  group_by(state) |>
  summarise(
    total_firms = sum(n_firms, na.rm = TRUE),
    ai_firms = sum(n_firms[used_tech == "Used"], na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    state = str_pad(state, 2, pad = "0"),
    ai_rate = if_else(total_firms > 0, ai_firms / total_firms, NA_real_)
  ) |>
  filter(total_firms > 0)

# Fetch map boundaries and shift AK/HI to the bottom
us_map <- tigris::states(cb = TRUE, resolution = "20m", class = "sf") |>
  filter(STUSPS != "PR") |>
  tigris::shift_geometry() |>
  mutate(GEOID = str_pad(GEOID, 2, pad = "0"))

map_data <- us_map |>
  left_join(state_ai_adoption, by = c("GEOID" = "state"))


ggplot(map_data) +
  geom_sf(aes(fill = ai_rate), color = "white", linewidth = 0.2) +
  scale_fill_distiller(
    palette = "OrRd",
    direction = 1,
    labels = scales::percent_format(accuracy = 1),
    na.value = "grey95"
  ) +
  labs(
    title = "Where are the AI Hubs?",
    subtitle = "AI Adoption Rate by State in 2018",
    fill = "Adoption\nRate",
    caption = "Source: 2018 Annual Business Survey (ABS)"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "right"
  )
```




## Wage Gaps and Premium

```{r, data_prep_cs}
tech_2018_summary <- tech_2018_clean |>
  filter(
    !is.na(used_tech),
    MOTUSETECH_LABEL == "All firms",
    IMPACTWF_U_LABEL == "All firms"    ) |>
  group_by(year, tech_type, used_tech, use_level) |>
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop") |>
  mutate(
    emp_per_firm = if_else(total_firms > 0, total_emp / total_firms, NA_real_),
    avg_wage     = if_else(total_emp > 0, total_payroll / total_emp, NA_real_))

industry_tech_summary <- tech_2018_clean |> 
  filter(
    !is.na(used_tech), 
    MOTUSETECH_LABEL == "All firms",   
    IMPACTWF_U_LABEL == "All firms"   ) |> 
  group_by(year, naics_label, tech_type, used_tech, use_level) |> 
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop"
  ) |> 
  mutate(
    emp_per_firm = if_else(total_firms > 0, total_emp / total_firms, NA_real_),
    avg_wage     = if_else(total_emp > 0, total_payroll / total_emp, NA_real_)
  )

ai_industry_wages <- industry_tech_summary |>
  filter(
    tech_type == "artificial intelligence",
    used_tech %in% c("Did not use", "Used")
  ) |>
  group_by(naics_label, used_tech) |>
  summarise(
    total_emp = sum(total_emp, na.rm = TRUE),
    total_payroll = sum(total_payroll, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(avg_wage = if_else(total_emp > 0, total_payroll / total_emp, NA_real_)) |>
  filter(total_emp > 0) 

# one row per industry with both groups side-by-side
wage_wide <- ai_industry_wages |>
  filter(used_tech %in% c("Did not use", "Used")) |>
  group_by(naics_label) |>
  filter(n() == 2) |>   # only industries where BOTH groups exist
  ungroup() |>
  pivot_wider(
    id_cols    = naics_label,
    names_from = used_tech,
    values_from = c(avg_wage, total_emp, total_payroll)) |>
  rename(
    wage_no_ai = `avg_wage_Did not use`,
    wage_ai    = `avg_wage_Used`,
    emp_no_ai  = `total_emp_Did not use`,
    emp_ai     = `total_emp_Used`) 

 wage_wide <- wage_wide |>
  mutate(
    wage_gap = wage_ai - wage_no_ai,
    premium_pct = if_else(wage_no_ai > 0, wage_gap / wage_no_ai, NA_real_),
    total_emp = emp_no_ai + emp_ai,
    ai_emp_share = if_else(total_emp > 0, emp_ai / total_emp, NA_real_)
  )

```


```{r, plots_cs}
##  The AI Wage Gap by Industry
dumbbell_long <- wage_wide |>
  pivot_longer(
    cols = c(wage_no_ai, wage_ai),
    names_to = "group",
    values_to = "avg_wage") |>
  mutate(
    group = if_else(group == "wage_ai", "AI Adopters", "Non-Adopters"),
    naics_label = reorder(naics_label, wage_gap))

ggplot() +
  geom_segment(
    data = wage_wide |> mutate(naics_label = reorder(naics_label, wage_gap)),
    aes(x = wage_no_ai, xend = wage_ai,
        y = naics_label, yend = naics_label),
    color = "grey75", linewidth = 1.3) +
  geom_point(
    data = dumbbell_long,
    aes(x = avg_wage, y = naics_label, color = group),
    size = 4) +
  scale_color_viridis_d(option = "D") +
  scale_x_continuous(labels = scales::label_dollar(suffix = "K")) +
  labs(
    title = "The AI Wage Gap by Industry",
    subtitle = "AI adopters vs. non-adopters (2018)",
    x = "Avg. Annual Payroll per Employee (Thousands USD)",
    y = NULL,
    color = NULL,
    caption = "Source: 2018 ABS") +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8))



## AI Wage Premium: Who Benefits Most?

wage_wide |>
  mutate(
    naics_label  = reorder(naics_label, premium_pct),
    premium_sign = if_else(premium_pct >= 0, "Positive", "Negative")
  ) |>
  ggplot(aes(x = premium_pct, y = naics_label, fill = premium_sign)) +
  geom_col(show.legend = FALSE, width = 0.75) +
  geom_vline(xintercept = 0, color = "grey40", linewidth = 0.5) +
  # Label the bars with the % value
  geom_text(
    aes(
      label = scales::percent(premium_pct, accuracy = 1),
      hjust = if_else(premium_pct >= 0, -0.15, 1.15)
    ),
    size = 3, color = "grey20"
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.08, 0.10))
  ) +
  scale_fill_manual(values = c("Positive" = "#2d6a4f", "Negative" = "#c1121f")) +
  labs(
    title    = "AI Wage Premium: Who Benefits Most?",
    subtitle = "% difference in avg. pay between AI and non-AI firms, by industry (2018)",
    x        = "Wage Premium (%)",
    y        = NULL,
    caption  = "Source: 2018 Annual Business Survey (ABS)\nNote: Food Services and Information sectors show the largest wage premiums for AI-adopting firms."
  )


## Employment Scale vs. Wage Premium
ggplot(
  wage_wide |> mutate(premium_sign = if_else(premium_pct >= 0, "Positive", "Negative")),
  aes(
    x = ai_emp_share,
    y = premium_pct,
    size = total_emp,
    color = premium_sign,
    label = naics_label
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.5) +
  geom_vline(xintercept = 0.15, linetype = "dotted", color = "grey60", linewidth = 0.5) +
  geom_point(alpha = 0.65) +
  geom_text(
    size = 3,
    color = "grey20") +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(-0.15, 0.75)
  ) +
  scale_size_continuous(
    range = c(3, 16),
    labels = scales::label_number(scale = 1e-6, suffix = "M", accuracy = 1),
    name = "Total Employees"
  ) +
  scale_color_manual(
    values = c("Positive" = "#2d6a4f", "Negative" = "#c1121f"),
    guide = "none"
  ) +
  labs(
    title = "Where Does AI Create the Most Value?",
    subtitle = "Share of industry workforce at AI firms vs. wage premium",
    x = "Share of Industry Employees at AI Firms",
    y = "AI Wage Premium",
    caption = "Source: 2018 Annual Business Survey (ABS)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    plot.caption = element_text(hjust = 0, color = "grey40", size = 8)
  )
```


## Patterns within Industries


## AI Adoption Spatial

> Note: industry sectors with more than 10 missing states were excluded from this analysis. These sectors are Agriculture, Mining, and Utilities. The remaining sectors provide a more reliable look at AI adoption trends. AI adoption rates are rendered to two decimal places to capture variations in sectors where adoption is emerging but still below 1%.

## Takeaways

> **AI adoption in 2018 was driven by firm scale, not industry wealth or geography.**

  - Adoption did not follow the money

Management led AI adoption at 7.7% (Figure 1), yet ranked near the bottom in total 2017
revenue (Figure 4). Wholesale Trade and Manufacturing — the two largest industries by revenue —were mid-tier adopters at best. Knowledge-intensive sectors moved first because AI fits their work, not because they had the deepest pockets.

  - The wage premium is real — and largest where adoption is rarest

AI adopters paid more at every firm size (Figure 5). But the premium was widest in sectors
where adoption was still uncommon: Food Services led at 86%, followed by Health Care (68%) and Wholesale Trade (67%) — all low-to-mid adoption industries (Figure 6). When most peers adopt, the competitive advantage narrows; Finance and Insurance, a top adopter, showed only 24%. The bubble chart (Figure 7) confirms this: industries with the highest premiums tend to have a smaller share of workers actually at AI firms, meaning early movers are still rare enough to capture outsized returns.

  - Geography is a proxy for industrial composition

California dominates the adoption map (Figure 8), but the random forest assigns **State
near-zero permutation importance** once firm characteristics are known (Figure 10). The
geographic pattern reflects where Management, Information, and Professional Services firms
concentrate — not something intrinsic to location.

  - Payroll growth favoured AI-adopting industries — with a COVID caveat

The top four AI adopters grew payroll steadily from 2017 to 2021 (Figure 2). Lower-adoption sectors, particularly Agriculture and Arts & Recreation, showed sharp swings around 2019–2020 (Figure 3) — a COVID artifact driven by in-person service disruption, not a structural signal. AI-adopting industries' relative stability over the same period may partly reflect the resilience of remote-compatible, technology-augmented operations.

  - *The forest confirmed scale as the dominant factor* 

The single decision tree (Figure 10) splits first on **number of firms ≥ 558**, then on
**employees per firm ≥ 61** — a profile that maps directly onto the concentrated,
knowledge-intensive industries at the top of Figure 1. Adding 2017 panel variables — revenue per firm and average industry wage — cut OOB prediction error from 5.3% to 3.1%, the largest accuracy gain in the analysis. Pre-adoption financial context carries genuine signal that cross-sectional 2018 data alone cannot capture.

The Gini vs. permutation comparison (Figure 9) shows that Gini inflated the apparent importance of continuous size variables. Permutation importance redistributed credit more
evenly and elevated employees-per-firm — suggesting AI adoption reflects a single latent
dimension of firm scale rather than any one metric in isolation.


This analysis is observational. Whether AI adoption *caused* the subsequent payroll divergence, or whether the same organizational characteristics that predict adoption also predict payroll growth, cannot be resolved with this data — and remains the natural next step.


## Appendix

### Machine Learning to predict AI Usage

After doing my EDA in the previous document, I was interested in whether we can predict which firms adopt AI. Thus, I decided to run a Random Forest.

#### Random Forest

The random forest algorithm extends bagging by combining both bagging and feature randomness to create an uncorrelated forest of decision trees. Decision trees work by asking a series of questions that split the data into branches. They are trained using the Classification and Regression Tree (CART) algorithm, with metrics like Gini impurity, information gain, or MSE (mean std. errors) used to evaluate split quality.

#### Building the Forest

While initial results using Gini-based importance suggested that firm size variables overwhelmingly drive AI adoption, this measure is known to be biased toward continuous predictors. Using permutation importance — which evaluates the impact of each variable on model accuracy — provides a more reliable ranking. The results confirm that firm scale remains the primary predictor of AI adoption, followed by workforce intensity and industry sector. In contrast, geographic location (state) contributes no predictive power once firm characteristics are considered.

Because firm size variables are highly correlated, permutation importance likely distributes predictive contribution across them, suggesting that AI adoption is driven by an underlying latent factor of firm scale rather than any single metric.


```{r forest_code}

tidymodels_prefer()

# 2017 industry financials (pre-adoption context)
industry_financials <- cs_panel_clean |>
  filter(year == 2017) |>
  group_by(naics_label) |>
  summarise(
    revenue_per_firm = mean(revenue_per_firm, na.rm = TRUE),
    avg_wage_panel   = mean(avg_wage,         na.rm = TRUE),
    .groups = "drop")

#  9-predictor dataset
rf_data_9 <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    used_tech %in% c("Used", "Did not use"),
    !is.na(n_emp), !is.na(payroll), !is.na(n_firms),
    n_emp > 0, payroll > 0, n_firms > 0,
    naics_sector != "Total"
  ) |>
  mutate(
    ai_adopted   = as.factor(if_else(used_tech == "Used", "Yes", "No")),
    avg_wage     = payroll / n_emp,
    emp_per_firm = n_emp / n_firms,
    naics_label  = as.factor(naics_label),
    state        = as.factor(state)
  ) |>
  left_join(industry_financials, by = "naics_label") |>
  select(
    ai_adopted,
    # 7-predictor core
    n_emp, n_firms, avg_wage, emp_per_firm, payroll, naics_label, state,
    # +2 from panel
    revenue_per_firm, avg_wage_panel
  ) |>
  drop_na()

# 7-predictor subset
rf_data_7 <- rf_data_9 |>
  select(-revenue_per_firm, -avg_wage_panel)

cat("Rows in both datasets:", nrow(rf_data_7), "(same rows, different columns)\n")
cat("7-predictor columns:", ncol(rf_data_7) - 1, "\n")
cat("9-predictor columns:", ncol(rf_data_9) - 1, "\n")


#Shared label map
var_labels <- c(
  "n_emp"            = "# Employees",
  "n_firms"          = "# Firms",
  "avg_wage"         = "Avg Wage (2018)",
  "emp_per_firm"     = "Employees / Firm",
  "payroll"          = "Total Payroll",
  "naics_label"      = "Industry Sector",
  "state"            = "State",
  "revenue_per_firm" = "Revenue / Firm (2017)",
  "avg_wage_panel"   = "Avg Industry Wage (2017)"
)


#  Forest specs
# settings: 1000 trees, min_n = 2, sqrt(p) predictors tried per split -- AKA random forest

make_spec <- function(importance_type) {
  rand_forest() |>
    set_mode("classification") |>
    set_engine("ranger") |>
    set_args(
      mtry        = NULL,       
      trees       = 1000,
      min_n       = 2,
      probability = FALSE,
      importance  = importance_type
    )
}

spec_gini  <- make_spec("impurity")    # Gini / mean decrease in impurity
spec_perm  <- make_spec("permutation") # Permutation importance


# Fit all four forests
set.seed(212)
forest_7_gini <- spec_gini |> fit(ai_adopted ~ ., data = rf_data_7)
forest_7_perm <- spec_perm |> fit(ai_adopted ~ ., data = rf_data_7)

set.seed(212)
forest_9_gini <- spec_gini |> fit(ai_adopted ~ ., data = rf_data_9)
forest_9_perm <- spec_perm |> fit(ai_adopted ~ ., data = rf_data_9)


#  OOB error comparison table
oob_summary <- tibble(
  Model      = c("7 predictors", "7 predictors", "9 predictors", "9 predictors"),
  Importance = c("Gini", "Permutation", "Gini", "Permutation"),
  OOB_Error  = c(
    forest_7_gini |> extract_fit_engine() |> pluck("prediction.error"),
    forest_7_perm |> extract_fit_engine() |> pluck("prediction.error"),
    forest_9_gini |> extract_fit_engine() |> pluck("prediction.error"),
    forest_9_perm |> extract_fit_engine() |> pluck("prediction.error"))) |>
  mutate(OOB_Accuracy = 1 - OOB_Error)

print(oob_summary)

# Visual: OOB error side-by-side
oob_summary |>
  ggplot(aes(x = Model, y = OOB_Error, fill = Importance)) +
  geom_col(position = "dodge", width = 0.55) +
  geom_text(
    aes(label = scales::percent(OOB_Error, accuracy = 0.1)),
    position = position_dodge(width = 0.55),
    vjust = -0.5, size = 3.5
  ) +
  scale_fill_manual(values = c("Gini" = "#2d6a4f", "Permutation" = "#f4a582")) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.1))) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "OOB Prediction Error: 7 vs 9 Predictors",
    subtitle = "Lower is better — error rate on out-of-bag samples",
    x        = NULL, y = "OOB Error Rate", fill = "Importance method",
    caption  = "Source: 2018 Annual Business Survey + 2017 Economic Census"
  )


# Importance extraction helper
get_importance <- function(forest, method_label, n_pred_label) {
  forest |>
    extract_fit_engine() |>
    pluck("variable.importance") |>
    enframe(name = "Variable", value = "Importance") |>
    mutate(
      Variable   = recode(Variable, !!!var_labels),
      Method     = method_label,
      Predictors = n_pred_label,
      # Normalise within each forest so Gini and Permutation scales are comparable
      Importance = Importance / max(abs(Importance))
    )
}

importance_all <- bind_rows(
  get_importance(forest_7_gini, "Gini",        "7 predictors"),
  get_importance(forest_7_perm, "Permutation",  "7 predictors"),
  get_importance(forest_9_gini, "Gini",        "9 predictors"),
  get_importance(forest_9_perm, "Permutation",  "9 predictors")
)


#  Importance comparison: Gini vs Permutation (within same predictor set)
plot_importance_comparison <- function(data, title_suffix) {
  data |>
    select(Variable, Method, Importance) |>
    pivot_wider(names_from = Method, values_from = Importance) |>
    mutate(Variable = fct_reorder(Variable, Gini)) |>
    ggplot(aes(y = Variable)) +
    geom_segment(
      aes(x = Permutation, xend = Gini, yend = Variable),
      color = "grey70", linewidth = 0.6
    ) +
    geom_point(aes(x = Gini,        color = "Gini"),        size = 3) +
    geom_point(aes(x = Permutation, color = "Permutation"), size = 3) +
    scale_color_manual(values = c("Gini" = "#2d6a4f", "Permutation" = "#e76f51")) +
    theme_minimal(base_size = 11) +
    labs(
      title    = glue::glue("Gini vs Permutation Importance — {title_suffix}"),
      subtitle = "Normalised 0–1 within each method. Dots far apart = Gini bias.",
      x = "Normalised importance", y = NULL, color = NULL,
      caption = "Source: 2018 Annual Business Survey"
    ) +
    theme(legend.position = "bottom")
}

p7 <- importance_all |>
  filter(Predictors == "7 predictors") |>
  plot_importance_comparison("7 predictors")

p9 <- importance_all |>
  filter(Predictors == "9 predictors") |>
  plot_importance_comparison("9 predictors")

p7 / p9   # patchwork: stacked


#  Faceted importance
importance_all |>
  group_by(Method, Predictors) |>
  mutate(Variable = fct_reorder(Variable, Importance)) |>
  ungroup() |>
  ggplot(aes(x = Importance, y = Variable, color = Method)) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_manual(values = c("Gini" = "#2d6a4f", "Permutation" = "#e76f51")) +
  facet_grid(Predictors ~ Method, scales = "free_y") +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold")) +
  labs(
    title    = "Variable Importance: All Four Models",
    subtitle = "Normalised importance — rows sorted independently per panel",
    x = "Normalised importance", y = NULL, color = NULL,
    caption = "Source: 2018 Annual Business Survey + 2017 Economic Census"
  )


# OOB Confusion matrices
# (using the permutation forests since they have more honest importance estimates)
cat("\n── Confusion matrix: 7-predictor forest (permutation) ──\n")
forest_7_perm |>
  extract_fit_engine() |>
  pluck("confusion.matrix") |>
  t() |>
  print()

cat("\n── Confusion matrix: 9-predictor forest (permutation) ──\n")
forest_9_perm |>
  extract_fit_engine() |>
  pluck("confusion.matrix") |>
  t() |>
  print()


# Single decision tree 
# The forest uses 1000 of these, each on a bootstrap sample with random feature
# subsets — individual trees are weak but their average is robust.

single_tree <- decision_tree() |>
  set_mode("classification") |>
  set_engine("rpart") |>
  set_args(cost_complexity = 0.02, tree_depth = 4) |>
  fit(
    ai_adopted ~ n_firms + emp_per_firm + avg_wage + payroll + n_emp,
    data = rf_data_7   # no state & naics_label to keep the tree readable
  )

single_tree |>
  extract_fit_engine() |>
  rpart.plot(
    type        = 4,
    extra       = 104,          # show % and n at each leaf
    box.palette = "GnRd",
    roundint    = FALSE,
    fallen.leaves = TRUE,      
    main = paste(
      "One example tree (depth 4, cost_complexity = 0.02)",
      "The random forest grows 1000 of these on bootstrapped samples",
      "Each split considers only sqrt(5) ≈ 2 randomly chosen predictors",
      sep = "\n"
    )
  )
```



