---
title: "David's Exploration"
format: 
  html:
    toc: true
    code-fold: true
---

## Motivations

In this document, I will be looking at the evolution of AI. Since we only have data for 2018 on AI-usage, I would look within those industries leading the AI race, how much have their payroll evolve. We excluded the *total* observations, as well as the *unknown* category.

## Introduction

```{r, setup, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(plotly)
library(dplyr)
library(tigris)
library(shiny)
library(leaflet)
library(sf)
library(stringr)
library(scales) 
library(ggrepel)
library(tidymodels)
# library(gam)
library(randomForest)
library(ranger)
library(vip)

source(here::here("data", "data_cleaning.R"))

## Theme Settings
theme_set(
  theme_minimal(base_size = 12) + 
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30"),
      plot.caption = element_text(hjust = 0, color = "grey40", size = 8),
      panel.grid.minor = element_blank()
    )
)

scale_use_level <- function(...) {
  scale_color_brewer(palette = "Dark2", name = NULL, ...)
}

scale_use_level_fill <- function(...) {
  scale_fill_brewer(palette = "Dark2", name = NULL, ...)
}

scale_use_level_fill_c <- function(...) {
  scale_fill_distiller(palette = "Dark2", direction = 1, name = NULL, ...)
}


## Things I have to fix: I did some renaming later on, so I need to replace all.
# cs_panel_clean <- cs_panel_clean |> 
#   rename(
#     n_firms = FIRMPDEMP,  ## done
#     n_emp   = EMP, ## same
#     payroll = payroll, ## done
#     revenue = revenue) ## done
```


## Who's the top industry using AI?

```{r top_ai}
ai_industry_leaders <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence", 
    !is.na(used_tech), 
    used_tech != "Unknown"
  ) |>
  group_by(naics_label) |>
  summarise(
    total_firms = sum(n_firms, na.rm = TRUE),
    ai_firms = sum(n_firms[used_tech == "Used"], na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(ai_rate = ai_firms / total_firms) |>
  filter(total_firms > 0) |>
  arrange(desc(ai_rate))

ggplot(ai_industry_leaders, aes(x = reorder(naics_label, ai_rate), y = ai_rate)) +
  geom_col(fill = "grey30") + # Standardizing color for non-categorical bars
  coord_flip() + 
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Which Industries Led the AI Race in 2018?",
    subtitle = "Percentage of firms reporting AI Use",
    x = NULL,
    y = "AI Adoption Rate"
  )
```

```{r}
tech_2018_summary <- tech_2018_clean |> 
  group_by(year, used_tech, use_level) |> 
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop"
  )

payroll_all <- cs_panel_clean |>
  group_by(year, naics_label) |>
  summarise(total_payroll = sum(payroll * 1000, na.rm = TRUE), .groups = "drop")

top_4_industries <- ai_industry_leaders |> slice_max(ai_rate, n = 4) |> pull(naics_label)
payroll_top4 <- payroll_all |> filter(naics_label %in% top_4_industries)

ggplot() +
  geom_line(data = payroll_all, 
            aes(x = year, y = total_payroll, group = naics_label), 
            color = "grey85", linewidth = 0.5, alpha = 0.6) +
  geom_line(data = payroll_top4, 
            aes(x = year, y = total_payroll, color = naics_label, group = naics_label), 
            linewidth = 1) +
  geom_point(data = payroll_top4, 
             aes(x = year, y = total_payroll, color = naics_label), 
             size = 1.5) +
  # Added accuracy = 1 right here to remove the .00
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-9, suffix = " B", accuracy = 1)) +
  scale_use_level() + 
  labs(
    title = "Payroll Evolution",
    subtitle = "Highlighting the 4 industries with the highest AI adoption rates in 2018",
    x = "Year",
    y = "Total Annual Payroll (Billions)",
    color = NULL,
    caption = "Healthcare is the top industry by total payroll, despite showing low AI adoption rates in 2018."
  ) + theme(legend.position = "top")
```
## ?? Revenue

```{r}
# 1. Filter for 2017 and aggregate revenue
revenue_2017 <- cs_panel_clean |>
  filter(year == 2017) |>
  group_by(naics_label) |>
  summarise(total_revenue = sum(revenue * 1000, na.rm = TRUE), .groups = "drop") |>
  # Filter out any lingering zeros just to be safe
  filter(total_revenue > 0) |>
  arrange(desc(total_revenue))

#  2017 snapshot (in Trillions)
ggplot(revenue_2017, aes(x = total_revenue, y = reorder(naics_label, total_revenue))) +
  geom_col(fill = "grey30", width = 0.9) + 
  scale_x_continuous(
    labels = scales::label_dollar(scale = 1e-12, suffix = "T", accuracy = 1),
    breaks = seq(0, 10e12, by = 1e12),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Total Industry Revenue in 2017",
    subtitle = "A snapshot of financial volume right before the 2018 AI Survey",
    x = "Total Revenue (Trillions)",
    y = NULL,
    caption = "Source: 2017 Economic Census"
  ) +  theme_minimal(base_size = 12) + 
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    plot.caption = element_text(hjust = 0, color = "grey40", size = 8),
    panel.grid.minor = element_blank())
```


## The AI Wage Premium

Does the AI wage premium holds constant as firms grow in size?

```{r}
## wage premium story
tech_2018_clean |>
  filter(tech_type == "artificial intelligence",
         used_tech != "Unknown",
         naics_sector != "Total", 
         payroll > 0,
         n_emp > 0) |>
  mutate(
    ai_used = if_else(used_tech == "Used", "AI Adopters", "Non-Adopters"),
    payroll_per_emp = payroll / n_emp) |> 
  ggplot(aes(x = n_emp, y = payroll_per_emp, color = ai_used)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth( se = FALSE, linewidth = 1.5) +  # method = "mgcv",
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(
    labels = scales::label_dollar(suffix = "K"), 
    breaks = seq(0, 300, by = 50)
  ) +  
  scale_use_level() +
  labs(
    title = "Does AI Pay?",
    subtitle = "Evidence of higher payroll standards among technology adopters across firm scales",
    x = "Number of Employees (Log Scale)",
    y = "Avg. Payroll per Employee (USD)",
    color = "Adoption Status",
    caption = "Source: 2018 Annual Business Survey (ABS) | Author: David Rios\nNote: Payroll standardized to thousands of dollars per employee. Trend lines calculated via local regression (LOESS)."
  ) +
  theme_minimal() + 
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8))
```


## AI Hubs in the US

```{r}
# AI Adoption Rate by State
state_ai_adoption <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    !is.na(used_tech),
    used_tech != "Unknown"
  ) |>
  group_by(state) |>
  summarise(
    total_firms = sum(n_firms, na.rm = TRUE),
    ai_firms = sum(n_firms[used_tech == "Used"], na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(ai_rate = ai_firms / total_firms) |>
  filter(total_firms > 0)

# Fetch map boundaries and shift AK/HI to the bottom
us_map <- tigris::states(cb = TRUE, resolution = "20m", class = "sf") |> 
  tigris::shift_geometry()

map_data <- us_map |>
  left_join(state_ai_adoption, by = c("GEOID" = "state"))
ggplot(map_data) +
  geom_sf(aes(fill = ai_rate), color = "white", linewidth = 0.2) +
  scale_fill_distiller(
    palette = "OrRd", 
    direction = 1, 
    labels = scales::percent_format(accuracy = 1),
    na.value = "grey95"
  ) +
  labs(
    title = "Where are the AI Hubs?",
    subtitle = "AI Adoption Rate by State in 2018",
    fill = "Adoption\nRate",
    caption = "Source: 2018 Annual Business Survey (ABS)"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "right")
```




## Wage Gaps and Premium

```{r data_preparation_cs}
tech_2018_summary <- tech_2018_clean |>
  filter(
    !is.na(used_tech),
    MOTUSETECH_LABEL == "All firms",
    IMPACTWF_U_LABEL == "All firms"    ) |>
  group_by(year, tech_type, used_tech, use_level) |>
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop") |>
  mutate(
    emp_per_firm = if_else(total_firms > 0, total_emp / total_firms, NA_real_),
    avg_wage     = if_else(total_emp > 0, total_payroll / total_emp, NA_real_))

industry_tech_summary <- tech_2018_clean |> 
  filter(
    !is.na(used_tech), 
    MOTUSETECH_LABEL == "All firms",   
    IMPACTWF_U_LABEL == "All firms"   ) |> 
  group_by(year, naics_label, tech_type, used_tech, use_level) |> 
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop"
  ) |> 
  mutate(
    emp_per_firm = if_else(total_firms > 0, total_emp / total_firms, NA_real_),
    avg_wage     = if_else(total_emp > 0, total_payroll / total_emp, NA_real_)
  )

ai_industry_wages <- industry_tech_summary |>
  filter(
    tech_type == "artificial intelligence",
    used_tech %in% c("Did not use", "Used")
  ) |>
  group_by(naics_label, used_tech) |>
  summarise(
    total_emp = sum(total_emp, na.rm = TRUE),
    total_payroll = sum(total_payroll, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(avg_wage = if_else(total_emp > 0, total_payroll / total_emp, NA_real_)) |>
  filter(total_emp > 0) 

# one row per industry with both groups side-by-side
wage_wide <- ai_industry_wages |>
  filter(used_tech %in% c("Did not use", "Used")) |>
  group_by(naics_label) |>
  filter(n() == 2) |>   # only industries where BOTH groups exist
  ungroup() |>
  pivot_wider(
    id_cols    = naics_label,
    names_from = used_tech,
    values_from = c(avg_wage, total_emp, total_payroll)) |>
  rename(
    wage_no_ai = `avg_wage_Did not use`,
    wage_ai    = `avg_wage_Used`,
    emp_no_ai  = `total_emp_Did not use`,
    emp_ai     = `total_emp_Used`) |>
  mutate(
    wage_gap     = wage_ai - wage_no_ai,
    premium_pct  = wage_gap / wage_no_ai,
    total_emp    = emp_no_ai + emp_ai,
    ai_emp_share = emp_ai / total_emp)
```


```{r plots_cs}
##  The AI Wage Gap by Industry
dumbbell_long <- wage_wide |>
  pivot_longer(
    cols      = c(wage_no_ai, wage_ai),
    names_to  = "group",
    values_to = "avg_wage") |>
  mutate(
    group       = if_else(group == "wage_ai", "AI Adopters", "Non-Adopters"),
    naics_label = reorder(naics_label, wage_gap))

ggplot() +
  geom_segment(
    data = wage_wide |> mutate(naics_label = reorder(naics_label, wage_gap)),
    aes(x = wage_no_ai, xend = wage_ai,
        y = naics_label,  yend = naics_label),
    color = "grey75", linewidth = 1.3) +
  geom_point(
    data = dumbbell_long,
    aes(x = avg_wage, y = naics_label, color = group),
    size = 4) +
  scale_use_level() +
  scale_x_continuous(labels = scales::label_dollar(suffix = "K")) +
  labs(
    title    = "The AI Wage Gap by Industry",
    subtitle = "Average annual payroll per employee: AI adopters vs. non-adopters (2018)",
    x        = "Avg. Annual Payroll per Employee (Thousands USD)",
    y        = NULL,
    color    = NULL,
    caption  = "Source: 2018 Annual Business Survey (ABS)\nRetail Trade is the only industry where AI adopters earn less on average — likely driven by logistics automation.") +
  theme(legend.position = "top")


## AI Wage Premium: Who Benefits Most?

wage_wide |>
  mutate(
    naics_label  = reorder(naics_label, premium_pct),
    premium_sign = if_else(premium_pct >= 0, "Positive", "Negative")
  ) |>
  ggplot(aes(x = premium_pct, y = naics_label, fill = premium_sign)) +
  geom_col(show.legend = FALSE, width = 0.75) +
  geom_vline(xintercept = 0, color = "grey40", linewidth = 0.5) +
  # Label the bars with the % value
  geom_text(
    aes(
      label = scales::percent(premium_pct, accuracy = 1),
      hjust = if_else(premium_pct >= 0, -0.15, 1.15)
    ),
    size = 3, color = "grey20"
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.08, 0.10))
  ) +
  scale_fill_manual(values = c("Positive" = "#2d6a4f", "Negative" = "#c1121f")) +
  labs(
    title    = "AI Wage Premium: Who Benefits Most?",
    subtitle = "% difference in avg. pay between AI and non-AI firms, by industry (2018)",
    x        = "Wage Premium (%)",
    y        = NULL,
    caption  = "Source: 2018 Annual Business Survey (ABS)\nNote: Food Services and Information sectors show the largest wage premiums for AI-adopting firms."
  )


## Employment Scale vs. Wage Premium
wage_wide |>
  mutate(premium_sign = if_else(premium_pct >= 0, "Positive", "Negative")) |>
  ggplot(aes(
    x     = ai_emp_share,
    y     = premium_pct,
    size  = total_emp,
    color = premium_sign,
    label = naics_label)) +
  geom_hline(yintercept = 0,    linetype = "dashed", color = "grey60", linewidth = 0.5) +
  geom_vline(xintercept = 0.15, linetype = "dotted", color = "grey60", linewidth = 0.5) +
  geom_point(alpha = 0.65) +
  geom_text_repel(size = 3, color = "grey20", max.overlaps  = 20,
                  segment.color = "grey70",
                  nudge_y = if_else(wage_wide$naics_label == "Construction", -0.09, 0),
                  nudge_x = case_when(
                    wage_wide$naics_label == "Real Estate"             ~  0.04,
                    wage_wide$naics_label == "Administrative Services" ~  0.04,
                    TRUE                                               ~  0))+
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.05, 0.05))) +
  scale_y_continuous(limits = c(-0.15, 0.75), labels = scales::percent_format(accuracy = 1)) +
  scale_size_continuous(
    range  = c(3, 16),
    labels = scales::label_number(scale = 1e-6, suffix = "M", accuracy = 1),
    name   = "Total Employees") +
  scale_color_manual(
    values = c("Positive" = "#2d6a4f", "Negative" = "#c1121f"),
    guide  = "none") +
  labs(title    = "Where Does AI Create the Most Value?",
    subtitle = "Share of industry workforce at AI firms vs. the wage premium they command",
    x        = "Share of Industry Employees at AI Firms",
    y        = "AI Wage Premium (%)",
    caption  = "Source: 2018 Annual Business Survey (ABS)\nBubble size = total industry employment. Dashed line marks 0% premium; dotted line marks 15% AI workforce share.") +
  theme(legend.position = "right") +
  theme_set()
```



### Company Size Panel Clean (2017-2021)

> While adoption alone signals a premium, intensity is the true driver. In sectors like Finance and Information, high implementation leads to the most significant wage increases.

## Patterns within Industries


## AI Adoption Spatial

> Note: industry sectors with more than 10 missing states were excluded from this analysis. These sectors are Agriculture, Mining, and Utilities. The remaining sectors provide a more reliable look at AI adoption trends. AI adoption rates are rendered to two decimal places to capture variations in sectors where adoption is emerging but still below 1%.

## AI Adoption Spatial

```{r, maps, eval=FALSE}
# --- PREPARATION: State Lookups ---
fips_to_name <- tibble(
  state_fips = c("01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", "38", "39", "40", "41", "42", "44", "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", "56"),
  state_name = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming")
)

# --- STEP 1: Geometry (Silent & Clean) ---
sf_use_s2(FALSE)
states_sf <- suppressMessages(suppressWarnings(
  st_as_sf(maps::map("state", plot = FALSE, fill = TRUE)) |>
    mutate(state_name = str_to_title(word(ID, 1, sep = ":"))) |>
    st_make_valid() |> 
    group_by(state_name) |> 
    summarise(geometry = st_union(geom), .groups = "drop") |>
    st_transform(4326)
))

# --- STEP 2: AI Rates (Summarized & Defensive) ---
ai_rates <- tech_2018_clean |>
  ungroup() |> 
  filter(tech_type == "artificial intelligence") |>
  mutate(naics_sector = as.character(naics_sector)) |>
  group_by(state, naics_sector) |>
  summarise(
    ai_adoption_rate = sum(n_firms[used_tech == "Used"], na.rm = TRUE) / 
      sum(n_firms, na.rm = TRUE),
    .groups = "drop"
  )

# --- STEP 3: Payroll (Matching by Cleaned Sector Names) ---
payroll_stats <- cs_panel_clean |>
  ungroup() |> 
  group_by(state, naics_sector) |>
  summarise(
    mean_payroll_per_emp = mean(payroll_per_emp, na.rm = TRUE), # Raw USD
    .groups = "drop"
  )

# --- STEP 4: Final Merge & Industry Filter ---
state_ai_spatial <- ai_rates |>
  left_join(payroll_stats, by = c("state", "naics_sector")) |>
  left_join(fips_to_name, by = c("state" = "state_fips")) |>
  filter(!is.na(state_name))

# Identify industries with enough data to show in the dropdown
valid_industries <- state_ai_spatial |>
  group_by(naics_sector) |>
  summarise(missing = sum(is.na(ai_adoption_rate)), .groups = "drop") |>
  filter(missing <= 10, !naics_sector %in% c("Unknown", "Total")) |>
  pull(naics_sector)

# --- STEP 5: Shiny App ---
ui <- fluidPage(
  titlePanel("US AI Adoption & Payroll by Sector (2018)"),
  sidebarLayout(
    sidebarPanel(
      selectInput("industry_choice", "Select Industry Sector:", 
                  choices = sort(valid_industries),
                  selected = "Professional Services"),
      hr(),
      helpText("Hover over states for metrics. Grey indicates insufficient data.")
    ),
    mainPanel(leafletOutput("map", height = "750px"))
  )
)

server <- function(input, output, session) {
  
  filtered_data <- reactive({
    req(input$industry_choice)
    state_ai_spatial |>
      filter(naics_sector == input$industry_choice) |>
      left_join(states_sf, by = "state_name") |>
      st_as_sf() |>
      filter(!st_is_empty(geometry))
  })
  
  output$map <- renderLeaflet({
    leaflet() |>
      addProviderTiles(providers$CartoDB.Positron) |>
      setView(lng = -98.58, lat = 39.82, zoom = 4)
  })
  
  observe({
    data <- filtered_data()
    req(nrow(data) > 0)
    
    pal <- colorNumeric("YlOrRd", 
                        domain = c(0, max(data$ai_adoption_rate * 100, na.rm = TRUE)), 
                        na.color = "#D3D3D3")
    
    labels <- sprintf(
      "<strong>%s</strong><br/>
       Sector: %s<br/>
       AI Adoption: <b>%s</b><br/>
       Payroll/Emp: <b>%s</b>",
      data$state_name, 
      data$naics_sector, 
      ifelse(is.na(data$ai_adoption_rate), 
             "Insufficient Data", 
             paste0(round(data$ai_adoption_rate * 100, 2), "%")),
      # Applied / 1000 fix here for raw USD to 'k' display
      ifelse(is.na(data$mean_payroll_per_emp), 
             "N/A", 
             paste0("$", formatC(data$mean_payroll_per_emp / 1000, format="f", digits=1, big.mark=","), "k"))
    ) |> lapply(htmltools::HTML)
    
    leafletProxy("map", data = data) |>
      clearShapes() |>
      addPolygons(
        fillColor = ~pal(ai_adoption_rate * 100),
        fillOpacity = 0.8, color = "white", weight = 1,
        highlightOptions = highlightOptions(weight = 3, color = "#666", bringToFront = TRUE),
        label = labels
      ) |>
      clearControls() |>
      addLegend("bottomright", pal = pal, values = ~ai_adoption_rate * 100,
                title = "AI Adoption (%)", labFormat = labelFormat(suffix = "%"),
                na.label = "No Data")
  })
}

shinyApp(ui, server)
```


## Random Forest



## Motivations

After doing my EDA in the previous document, I was interested in whether we  canpredict which firms adopt AI. Thus, I decided to run a Random Forest.

## Uploading Libraries

```{r, set_2, message=FALSE, warning=FALSE}
tidymodels_prefer()
```


```{r}
## ── 1. Prep the data ───────────────────────────────────────────────────────

rf_data <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    used_tech %in% c("Used", "Did not use"),  # drop Unknown
    !is.na(n_emp), !is.na(payroll), !is.na(n_firms),
    n_emp > 0, payroll > 0, n_firms > 0,
    naics_sector != "Total"
  ) |>
  mutate(
    ai_adopted  = as.factor(if_else(used_tech == "Used", "Yes", "No")),
    avg_wage    = payroll / n_emp,
    emp_per_firm = n_emp / n_firms,
    naics_label = as.factor(naics_label),
    state       = as.factor(state)
  ) |>
  select(ai_adopted, n_emp, n_firms, avg_wage, emp_per_firm, 
         payroll, naics_label, state) |>
  drop_na()


## ── 2. Model Specification ─────────────────────────────────────────────────

rf_spec <- rand_forest() |>
  set_mode("classification") |>
  set_engine("ranger") |>
  set_args(
    mtry        = NULL,   # defaults to sqrt(n_predictors)
    trees       = 1000,
    min_n       = 2,
    probability = FALSE,
    importance  = "impurity"  # Gini index for variable importance
  )


## ── 3. Fit the forest ──────────────────────────────────────────────────────

set.seed(253)

ai_forest <- rf_spec |>
  fit(ai_adopted ~ ., data = rf_data)


## ── 4. OOB Error & Confusion Matrix ───────────────────────────────────────

# Overall OOB error rate
ai_forest

#  labels map
var_labels <- c(
  "n_firms"      = "Number of Firms",
  "emp_per_firm" = "Employees per Firm",
  "naics_label"  = "Industry Sector",
  "n_emp"        = "Number of Employees",
  "avg_wage"     = "Average Wage",
  "payroll"      = "Total Payroll",
  "state"        = "State")

## ── 5. Variable Importance ─────────────────────────────────────────────────
ai_forest |>
  extract_fit_engine() |>
  pluck("variable.importance") |>
  enframe(name = "Variable", value = "Importance") |>        # tibble
  mutate(Variable = recode(Variable, !!!var_labels)) |>      # rename
  arrange(Importance) |>
  mutate(Variable = fct_inorder(Variable)) |>                # preserve order
  ggplot(aes(x = Importance, y = Variable)) +
  geom_point(size = 3) +
  labs(
    title    = "What Drives AI Adoption?",
    subtitle = "Variable importance from Random Forest (Gini impurity index)",
    x        = "Importance",
    y        = NULL,
    caption  = "Source: 2018 Annual Business Survey (ABS)"
  )


## ── 6. In-sample confusion matrix ─────────────────────────────────────────

ai_forest |>
  predict(new_data = rf_data) |>
  cbind(rf_data) |>
  conf_mat(
    truth    = ai_adopted,
    estimate = .pred_class
  )
```


## Merging with Company Size

2017 snapshot as pre-adoption context:


```{r}
# Pull 2017 industry-level financials from cs_panel_clean
# ONE row per industry (national average)
industry_financials <- cs_panel_clean |>
  filter(year == 2017) |>
  group_by(naics_label) |>
  summarise(
    revenue_per_firm = mean(revenue_per_firm, na.rm = TRUE),
    avg_wage_panel   = mean(avg_wage,         na.rm = TRUE),
    .groups = "drop"
  )

# Rebuild rf_data with the new variables
rf_data <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    used_tech %in% c("Used", "Did not use"),
    !is.na(n_emp), !is.na(payroll), !is.na(n_firms),
    n_emp > 0, payroll > 0, n_firms > 0,
    naics_sector != "Total"
  ) |>
  mutate(
    ai_adopted   = as.factor(if_else(used_tech == "Used", "Yes", "No")),
    avg_wage     = payroll / n_emp,
    emp_per_firm = n_emp / n_firms,
    naics_label  = as.factor(naics_label),
    state        = as.factor(state)
  ) |>
  left_join(industry_financials, by = "naics_label") |>   # ← merge here
  select(
    ai_adopted, n_emp, n_firms, avg_wage, emp_per_firm,
    payroll, naics_label, state,
    revenue_per_firm, avg_wage_panel                       # ← new vars
  ) |>
  drop_na()

# Update labels map to include new variables
var_labels <- c(
  "n_firms"          = "Number of Firms",
  "emp_per_firm"     = "Employees per Firm",
  "naics_label"      = "Industry Sector",
  "n_emp"            = "Number of Employees",
  "avg_wage"         = "Average Wage (2018)",
  "payroll"          = "Total Payroll",
  "state"            = "State",
  "revenue_per_firm" = "Revenue per Firm (2017)",
  "avg_wage_panel"   = "Avg. Industry Wage (2017)"
)
```


```{r}
# ── Check the merged data ──────────────────────────────────────────────────
glimpse(rf_data)
# Should now show 9 predictors + ai_adopted, with revenue_per_firm & avg_wage_panel

# ── Re-fit the forest with new data ───────────────────────────────────────
set.seed(253)
ai_forest <- rf_spec |> fit(ai_adopted ~ ., data = rf_data)

# ── 1. OOB Summary ────────────────────────────────────────────────────────
ai_forest  # check if OOB error changed with new vars

# ── 2. OOB Confusion Matrix ───────────────────────────────────────────────
ai_forest |>
  extract_fit_engine() |>
  pluck("confusion.matrix") |>
  t()

# ── 3. Variable Importance Plot ───────────────────────────────────────────
ai_forest |>
  extract_fit_engine() |>
  pluck("variable.importance") |>
  enframe(name = "Variable", value = "Importance") |>
  mutate(Variable = recode(Variable, !!!var_labels)) |>
  arrange(Importance) |>
  mutate(Variable = fct_inorder(Variable)) |>
  ggplot(aes(x = Importance, y = Variable)) +
  geom_point(size = 3) +
  labs(
    title    = "What Drives AI Adoption?",
    subtitle = "Variable importance from Random Forest (Gini impurity index)",
    x        = "Importance (Gini)",
    y        = NULL,
    caption  = "Source: 2018 Annual Business Survey + 2017 Economic Census"
  )

# ── 4. In-sample Confusion Matrix (visualized) ────────────────────────────
ai_forest |>
  predict(new_data = rf_data) |>
  cbind(rf_data) |>
  conf_mat(truth = ai_adopted, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "white", high = "#2d6a4f") +  # ← green gradient
  labs(
    title    = "In-Sample Confusion Matrix",
    subtitle = "Random Forest: perfect in-sample fit (expected for unpruned trees)",
    caption  = "Note: OOB error (3.09%) is the honest estimate of out-of-sample performance."
  )

# ── 5. Single Example Tree (from the forest) ──────────────────────────────
# Forests can't be plotted directly, so we build ONE tree for illustration
library(rpart)
library(rpart.plot)
single_tree <- decision_tree() |>
  set_mode("classification") |>
  set_engine("rpart") |>
  set_args(cost_complexity = 0.01, tree_depth = 3) |>  # ← shallower
  fit(ai_adopted ~ n_firms + emp_per_firm + avg_wage + 
                   revenue_per_firm + avg_wage_panel,   # ← drop state & naics_label
      data = rf_data)

single_tree |>
  extract_fit_engine() |>
  rpart.plot(
    type        = 4,
    extra       = 104,
    box.palette = "GnRd",
    roundint    = FALSE,    # ← silences the warning
    main        = "Example Decision Tree\n(Simplified — forest uses 1000 of these)"
  )

# ── 6. Classification Regions (2 key vars) ────────────────────────────────
# Visualize how the forest splits on the 2 most important vars
grid <- expand.grid(
  n_firms      = seq(min(rf_data$n_firms), max(rf_data$n_firms), length.out = 150),
  emp_per_firm = seq(min(rf_data$emp_per_firm), max(rf_data$emp_per_firm), length.out = 150)
) |>
  mutate(
    n_emp            = median(rf_data$n_emp),
    avg_wage         = median(rf_data$avg_wage),
    payroll          = median(rf_data$payroll),
    naics_label      = "Manufacturing",   # ← neutral mid-adoption industry
    state            = "06",              # ← California, high adoption state
    revenue_per_firm = median(rf_data$revenue_per_firm, na.rm = TRUE),
    avg_wage_panel   = median(rf_data$avg_wage_panel,   na.rm = TRUE)
  )

ai_forest |>
  predict(new_data = grid) |>
  cbind(grid) |>
  ggplot(aes(x = n_firms, y = emp_per_firm, fill = .pred_class)) +
  geom_tile(alpha = 0.8) +
  scale_fill_manual(values = c("No" = "#f4a582", "Yes" = "#2d6a4f")) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title    = "Forest Classification Regions",
    subtitle = "AI adoption boundary: firm count vs. firm size (Manufacturing, CA)",
    x        = "Number of Firms (log scale)",
    y        = "Employees per Firm (log scale)",
    fill     = "AI Adopted?",
    caption  = "Source: 2018 Annual Business Survey (ABS)\nOther variables held at median values."
  )

```
