---
title: "Katherine M."
format: 
  html: 
    toc: true
    toc-depth: 3
    embed-resources: true
    code-overflow: wrap
    page-layout: full
---

## Motivations

In this document, I primarily look at disaggregated AI usage, both across states
and industries. I exclude *total* and *unknown* values to focus on known data. 
Alongside my teammates, I only focus on observations from 2018 as this was the 
only year with data on AI usage. 

I also look at current (May 2026) job trends related to AI, to determine whether 
current trends are similar to what we see in our data. 

## Introduction

Loading packages, data, and optional color themes for graphs. 

```{r}
# loading necessary packages for data visualizations
library(ggplot2)
library(dplyr)
library(forcats)
library(here)
library(tidyverse)
library(downlit)
library(xml2)
library(paletteer)
library(plotly)
library(cdlTools)
library(tigris)
library(sf)

source(here::here("data", "data_cleaning.R"))

## Theme Settings

scale_colour_paletteer_d("khroma::BuRd")
scale_color_paletteer_d("khroma::BuRd")
scale_fill_paletteer_d("khroma::BuRd")
paletteer_d("khroma::BuRd")
```

```{r}
# cleaning datasets for my viz purposes

k_tech_2018_clean <- tech_2018_clean |> 
  filter(tech_type == "artificial intelligence") |> 
  filter(naics_sector != "Total") |> 
  filter(naics_sector != "Other") |> 
  filter(!is.na(use_level)) |> 
  filter(!is.na(n_firms))

k_tech_2018_clean <- k_tech_2018_clean |> 
  group_by(state, naics_sector) |>
  mutate(industry_total = sum(n_firms, na.rm = TRUE),
         pct_of_industry = (n_firms / sum(n_firms, na.rm = TRUE)) * 100) |>
  ungroup()

# loading new dataset for job cuts
job_cuts <- read.csv("/Users/katiemanuel/Desktop/spring 26/stat212/project-economics-and-ds/data/job_cuts.csv")

job_cuts <- job_cuts |> 
  mutate(job_cuts = na_if(job_cuts, "Not specified")) |> 
  mutate(job_cuts = as.integer(gsub(",", "", job_cuts))) |> 
  mutate(month_announced = factor(month_announced, levels = month.name))
```


# AI Usage by State

How does AI usage vary by state & industry?

```{r eval=FALSE}
# Bar Graph

k_tech_2018_clean |> 
  mutate(n_firms = n_firms/100000) |> 
  ggplot(aes(x = fct_reorder(state_label, n_firms), y = n_firms, fill = use_level)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Firms by State (in hundred-thousands)", y = "# of Firms with High Reported AI Usage",
       title = "How Does AI Usage Vary Across the U.S.?",
       fill = "AI Usage Level",
       subtitle = "AI Usage by State in 2018",
       caption = "Source: Annual Business Survey (ABSTCB)") +
  scale_fill_paletteer_d("khroma::BuRd") +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8)) +
  coord_flip()
```

```{r}
# Spatial Visualization

# Spatial Viz: Top States Who "Did Not Use" AI
k_tech_2018_clean |> 
  # Grouping for total amounts. 
  group_by(state) |>
  mutate(overall_state_total = sum(n_firms, na.rm = TRUE)) |>
  ungroup() |>
  group_by(state, naics_sector) |>
  mutate(
    state_industry_total = sum(n_firms, na.rm = TRUE),
    pct_of_industry = (n_firms / state_industry_total) * 100) |>
  ungroup() |> 
  
  # Filtering to clean data. 
  mutate(state_abbr = cdlTools::fips(state, to = "Abbreviation")) |>
  filter(tech_type == "artificial intelligence") |> 
  filter(naics_sector != "Total") |> 
  filter(naics_sector != "Other") |> 
  filter(!is.na(use_level)) |> 
  filter(!is.na(n_firms)) |> 
  filter(use_level == "Did not use") |> 
  
  # Collapsing industries & preparing for plotly map. 
  group_by(state_abbr, state_label) |> 
  summarize(
    state_non_users = sum(n_firms, na.rm = TRUE),
    state_total_all_firms = first(overall_state_total),
    nonusers_to_allfirms = (state_non_users/state_total_all_firms), 
    industry_breakdown = paste(naics_label, ": ", round(pct_of_industry, 1), "%", collapse = "<br>"), .groups = "drop") |> 
  mutate(
    hover_text = paste(
      "<b>State:</b>", state_label,
      "<br><b>Total 'Did Not Use' Firms:</b>", scales::comma(state_non_users),
      "<br><b>Total Firms in State:</b>", scales::comma(state_total_all_firms),
      "<br><b>No-Use Firms/Total Firms in State:</b>", scales::comma(nonusers_to_allfirms),
      "<br><br><b>--- % of Non-Users by Industry ---</b><br>",
      industry_breakdown)) |> 
  plot_ly(
    type = "choropleth",
    locations = ~state_abbr,       
    locationmode = "USA-states",   
    z = ~nonusers_to_allfirms,          
    text = ~hover_text,            
    hoverinfo = "text",            
    colorscale = "Tropic",
    colorbar = list(title = "No-Use/Total Firms:")) |>
  layout(
    title = list(
      text = "<b>Firms That 'Did Not Use' AI by State and Industry</b><br><span style='color: grey; font-size: 14px;'>Across states, the vast majority of firms did not use AI.</span>",
      font = list(size = 16)),
    annotations = list(
      list(
        x = 0,               
        y = -0.1,            
        text = "Source: Annual Business Survey (ABSTCB)",
        showarrow = FALSE,
        xref = "paper",      
        yref = "paper",
        xanchor = "left",
        yanchor = "top",
        font = list(color = "grey", size = 8))),
    geo = list(
      scope = "usa",              
      projection = list(type = "albers usa"),
      showlakes = TRUE, 
      lakecolor = toRGB("white")))

# Spatial Viz: Top States With the Most Moderate AI Usage ("Low Use" and "Moderate Use" Combined)

k_tech_2018_clean |> 
  # Grouping for total amounts. 
  group_by(state) |>
  mutate(overall_state_total = sum(n_firms, na.rm = TRUE)) |>
  ungroup() |>
  group_by(state, naics_sector) |>
  mutate(
    state_industry_total = sum(n_firms, na.rm = TRUE),
    pct_of_industry = (n_firms / state_industry_total) * 100) |>
  ungroup() |> 
  
  # Filtering to clean data. 
  mutate(state_abbr = cdlTools::fips(state, to = "Abbreviation")) |>
  filter(tech_type == "artificial intelligence") |> 
  filter(naics_sector != "Total") |> 
  filter(naics_sector != "Other") |> 
  filter(!is.na(use_level)) |> 
  filter(!is.na(n_firms)) |> 
  filter(use_level %in% c("Moderate use", "Low use")) |> 
  
  # Collapsing industries & preparing for plotly map. 
  group_by(state_abbr, state_label) |> 
  summarize(
    state_non_users = sum(n_firms, na.rm = TRUE),
    state_total_all_firms = first(overall_state_total),
    nonusers_to_allfirms = (state_non_users/state_total_all_firms), 
    industry_breakdown = paste(naics_label, ": ", round(pct_of_industry, 1), "%", collapse = "<br>"), .groups = "drop") |> 
  mutate(
    hover_text = paste(
      "<b>State:</b>", state_label,
      "<br><b>Total 'Low/Moderate Use' Firms:</b>", scales::comma(state_non_users),
      "<br><b>Total Firms in State:</b>", scales::comma(state_total_all_firms),
      "<br><b>No-Use Firms/Total Firms in State:</b>", scales::comma(nonusers_to_allfirms),
      "<br><br><b>--- % of 'Some'-Users by Industry ---</b><br>",
      industry_breakdown)) |> 
  plot_ly(
    type = "choropleth",
    locations = ~state_abbr,       
    locationmode = "USA-states",   
    z = ~nonusers_to_allfirms,          
    text = ~hover_text,            
    hoverinfo = "text",            
    colorscale = "Tropic",
    colorbar = list(title = "Some-Use/Total Firms:")) |>
  layout(
    title = list(
      text = "<b>Firms That Reported 'Low/Moderate' AI Use by State and Industry</b><br><span style='color: grey; font-size: 14px;'>Across states, the vast majority of firms did not use AI.</span>",
      font = list(size = 16)),
    annotations = list(
      list(
        x = 0,               
        y = -0.1,            
        text = "Source: Annual Business Survey (ABSTCB)",
        showarrow = FALSE,
        xref = "paper",      
        yref = "paper",
        xanchor = "left",
        yanchor = "top",
        font = list(color = "grey", size = 8))),
    geo = list(
      scope = "usa",              
      projection = list(type = "albers usa"),
      showlakes = TRUE, 
      lakecolor = toRGB("white")))

# Spatial Viz: Top States With the Highest AI Usage 

k_tech_2018_clean |> 
  # Grouping for total amounts. 
  group_by(state) |>
  mutate(overall_state_total = sum(n_firms, na.rm = TRUE)) |>
  ungroup() |>
  group_by(state, naics_sector) |>
  mutate(
    state_industry_total = sum(n_firms, na.rm = TRUE),
    pct_of_industry = (n_firms / state_industry_total) * 100) |>
  ungroup() |> 
  
  # Filtering to clean data. 
  mutate(state_abbr = cdlTools::fips(state, to = "Abbreviation")) |>
  filter(tech_type == "artificial intelligence") |> 
  filter(naics_sector != "Total") |> 
  filter(naics_sector != "Other") |> 
  filter(!is.na(use_level)) |> 
  filter(!is.na(n_firms)) |> 
  filter(use_level == "High use") |> 
  
  # Collapsing industries & preparing for plotly map. 
  group_by(state_abbr, state_label) |> 
  summarize(
    state_non_users = sum(n_firms, na.rm = TRUE),
    state_total_all_firms = first(overall_state_total),
    nonusers_to_allfirms = (state_non_users/state_total_all_firms), 
    industry_breakdown = paste(naics_label, ": ", round(pct_of_industry, 1), "%", collapse = "<br>"), .groups = "drop") |> 
  mutate(
    hover_text = paste(
      "<b>State:</b>", state_label,
      "<br><b>Total 'High Use' Firms:</b>", scales::comma(state_non_users),
      "<br><b>Total Firms in State:</b>", scales::comma(state_total_all_firms),
      "<br><b>High-Use Firms/Total Firms in State:</b>", scales::comma(nonusers_to_allfirms),
      "<br><br><b>--- % of High-Users by Industry ---</b><br>",
      industry_breakdown)) |> 
  plot_ly(
    type = "choropleth",
    locations = ~state_abbr,       
    locationmode = "USA-states",   
    z = ~nonusers_to_allfirms,          
    text = ~hover_text,            
    hoverinfo = "text",            
    colorscale = "Tropic",
    colorbar = list(title = "High-Use/Total Firms:")) |>
  layout(
    title = list(
      text = "<b>Firms That Reported 'High' AI Use by State and Industry</b><br><span style='color: grey; font-size: 14px;'>Across states, the vast majority of firms did not use AI.</span>",
      font = list(size = 16)),
    annotations = list(
      list(
        x = 0,               
        y = -0.1,            
        text = "Source: Annual Business Survey (ABSTCB)",
        showarrow = FALSE,
        xref = "paper",      
        yref = "paper",
        xanchor = "left",
        yanchor = "top",
        font = list(color = "grey", size = 8))),
    geo = list(
      scope = "usa",              
      projection = list(type = "albers usa"),
      showlakes = TRUE, 
      lakecolor = toRGB("white")))
```

```{r}
# Which states have the LOWEST AI adoption rates? Non-Graph. 
k_tech_2018_clean |> 
    # Grouping for total amounts. 
  group_by(state) |>
  mutate(overall_state_total = sum(n_firms, na.rm = TRUE)) |>
  ungroup() |>
  group_by(state, naics_sector) |>
  mutate(
    state_industry_total = sum(n_firms, na.rm = TRUE),
    pct_of_industry = (n_firms / state_industry_total) * 100) |>
  ungroup() |> 
    # data cleaning 
  filter(tech_type == "artificial intelligence") |> 
  filter(naics_sector != "Total") |> 
  filter(naics_sector != "Other") |> 
  filter(!is.na(use_level)) |> 
  filter(!is.na(n_firms)) |> 
  filter(use_level == "Did not use") |> 
    # creating groups for summaries 
  group_by(state_label) |> 
  summarize(
    state_non_users = sum(n_firms, na.rm = TRUE),
    state_total_all_firms = first(overall_state_total),
    nonusers_to_allfirms = (state_non_users / state_total_all_firms)) |>
  arrange(desc(nonusers_to_allfirms)) |> 
  head(3)

# Which states have the HIGHEST AI adoption rates? (only counting "High use")
k_tech_2018_clean |> 
    # Grouping for total amounts. 
  group_by(state) |>
  mutate(overall_state_total = sum(n_firms, na.rm = TRUE)) |>
  ungroup() |>
  group_by(state, naics_sector) |>
  mutate(
    state_industry_total = sum(n_firms, na.rm = TRUE),
    pct_of_industry = (n_firms / state_industry_total) * 100) |>
  ungroup() |> 
    # data cleaning 
  filter(tech_type == "artificial intelligence") |> 
  filter(naics_sector != "Total") |> 
  filter(naics_sector != "Other") |> 
  filter(!is.na(use_level)) |> 
  filter(!is.na(n_firms)) |> 
  filter(use_level == "High use") |> 
    # creating groups for summaries 
  group_by(state_label) |> 
  summarize(
    state_high_users = sum(n_firms, na.rm = TRUE),
    state_total_all_firms = first(overall_state_total),
    highusers_to_allfirms = (state_high_users / state_total_all_firms)) |>
  arrange(desc(highusers_to_allfirms)) |> 
  head(3)

# Which states have the most MODERATE AI adoption rates? (only counting "Low use" and "Moderate Use")
k_tech_2018_clean |> 
    # Grouping for total amounts. 
  group_by(state) |>
  mutate(overall_state_total = sum(n_firms, na.rm = TRUE)) |>
  ungroup() |>
  group_by(state, naics_sector) |>
  mutate(
    state_industry_total = sum(n_firms, na.rm = TRUE),
    pct_of_industry = (n_firms / state_industry_total) * 100) |>
  ungroup() |> 
    # data cleaning 
  filter(tech_type == "artificial intelligence") |> 
  filter(naics_sector != "Total") |> 
  filter(naics_sector != "Other") |> 
  filter(!is.na(use_level)) |> 
  filter(!is.na(n_firms)) |> 
  filter(use_level %in% c("Moderate use", "Low use")) |> 
    # creating groups for summaries 
  group_by(state_label) |> 
  summarize(
    state_some_users = sum(n_firms, na.rm = TRUE),
    state_total_all_firms = first(overall_state_total),
    someusers_to_allfirms = (state_some_users / state_total_all_firms)) |>
  arrange(desc(someusers_to_allfirms)) |> 
  head(3)

# Confirming usage levels
unique(k_tech_2018_clean$use_level)
```

# AI Usage by Industry

How does AI usage vary by industry?

```{r eval=FALSE}
# high AI usage by industry from 2017-2021
tech_2018_clean |> 
  filter(tech_type == "artificial intelligence") |> 
  filter(naics_sector != "Total") |> 
  filter(naics_sector != "Other") |> 
  filter(!is.na(use_level)) |> 
  mutate(n_firms = n_firms/100000) |> 
  ggplot(aes(x = naics_label, y = n_firms, fill = use_level)) +
  geom_col() + 
  theme(axis.text.x = element_blank()) +
  labs(x = "Firms by Industry", 
       y = "# of Firms with High Reported AI Usage",
       title = "AI Usage by Industry (2017-2021)", 
       fill = "AI Usage") +
  theme_minimal() +
  scale_fill_paletteer_d("khroma::BuRd") +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8)) +
  coord_flip()
```

# Wages By Industry

What Does Average Wage Look Like Across Industries?

```{r}
# Total Industry Scatterplot, with Key Industries Marked Separately. 
cs_panel_clean |> 
  filter(year == 2018) |> 
  mutate(high_wage = if_else(
    naics_label %in% c("Finance and Insurance", "Information", "Management"), 
    naics_label, 
    "Other")) |>
  group_by(naics_label) |> 
  ggplot(aes(x = n_emp, y = avg_wage, color = high_wage)) +
  geom_point() +
  theme_minimal() +
  scale_color_paletteer_d("khroma::BuRd") +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8)) +
    labs(x = "Number of Employees", 
       y = "Average Wages Across States",
       title = "Which Industries Had the Highest Wages in 2018?", 
       color = "Industry", 
       subtitle = "Number of Employees and Average Wages in 2018",
       caption = "Source: Annual Business Survey (ABSCS)")

# Scatterplots separated by Industry. 
cs_panel_clean |> 
  filter(year == 2018) |> 
  group_by(naics_label) |> 
  ggplot(aes(x = n_emp, y = avg_wage)) +
  geom_point() +
  theme_minimal() +
  facet_wrap(~naics_label)
```

# Tech Hubs in the U.S.

Where are the major tech hubs in the U.S.? 

Note: Data taken from outside sources: https://online.king.edu/infographics/technology-hubs/
Plotted with R. 

```{r}
us_states <- states(cb = TRUE, resolution = "20m") |> 
  shift_geometry()

tech_hub_data <- us_states |>
  left_join(tech_2018_clean, by = c("NAME" = "state_label")) |>
  mutate(
    map_fill = if_else(
      NAME %in% c("California", "Washington", "Colorado", "Texas", 
                          "Massachusetts", "New York", "Tennessee"), 
      NAME,
      "Other"))

ggplot(tech_hub_data) +
  geom_sf(aes(fill = map_fill), color = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c(
      # dark blue are established hubs, light blue is up and coming
      "California"    = "#4393C3FF", 
      "Massachusetts" = "#4393C3FF", 
      "Washington"    = "#4393C3FF", 
      "Colorado"      = "#92C5DEFF",
      "Texas"         = "#92C5DEFF",
      "New York"      = "#4393C3FF",
      "Tennessee"     = "#92C5DEFF",
      "Other"         = "grey85"),
    breaks = c(
      "California", "Massachusetts", "New York", "Washington",
      "Colorado", "Tennessee", "Texas",
      "Other"),
    name = "States Containing AI Hubs") +
  theme_void() +
  theme(legend.position = "right") +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8)) +
  labs(title = "Tech Hubs Across the U.S.",
       subtitle = "Established hubs in dark blue. Up-and-coming hubs in light blue.", 
       caption = "Source: King University")
  
```

# Major Job Cuts Due to AI

How has AI impacted jobs in 2026?

```{r}
# heat map
job_cuts |> 
  filter(!is.na(job_cuts)) |> 
  ggplot(aes(x = month_announced, y = company, fill = job_cuts)) +
  geom_tile() +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8)) +
  labs(title = "2026 AI Job Cuts",
       subtitle = "More than 50K jobs have been/are currently estimated to be cut in 2026.",
       x = "Month Announced",
       y = "Number of Job Cuts", 
       fill = "Company", 
       caption = "Source: King University")

# bar graph
BuRd <- colorRampPalette(paletteer_d("khroma::BuRd"))(23) 

job_cuts |> 
  filter(!is.na(job_cuts)) |> 
  ggplot(aes(x = month_announced, y = job_cuts, fill = company)) +
  geom_col() +
  theme_minimal() +
  scale_fill_manual(values = BuRd) +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8)) +
  labs(title = "2026 AI Job Cuts",
       subtitle = "More than 50K jobs have been/are currently estimated to be cut in 2026.",
       x = "Month Announced",
       y = "Number of Job Cuts", 
       fill = "Company", 
       caption = "Source: Reuters")
```

# Takeaways

