---
title: "Isauro's EDA"
format: 
  html: 
    toc: true
    toc-depth: 3
    embed-resources: true
    code-overflow: wrap
    page-layout: full
---

```{r}
# load ggplot2
library(ggplot2)
library(here)
library(tidyverse)
library(plotly)

source(here::here("data", "data_cleaning.R"))
```

```{r}
# visualization settings

## plot functions

### column plot (continuous)
colplotCont <- function(df, xvar, yvar, fillvar) {
  ggplot(data = df, aes(x = {{xvar}}, y = {{yvar}}, fill = {{fillvar}})) +
    geom_col() +
    scale_fill_brewer(palette = "PuBu",
                     direction = 1) +
    theme_classic() + 
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30"),
      plot.caption = element_text(hjust = 0, color = "grey40", size = 8))
}

### column plot (discrete)
colplotDisc <- function(df, xvar, yvar, fillvar) {
  ggplot(data = df, aes(x = {{xvar}}, y = {{yvar}}, fill = {{fillvar}})) +
    geom_col() +
    scale_fill_distiller(palette = "PuBu",
                     direction = 1) +
    theme_classic() + 
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30"),
      plot.caption = element_text(hjust = 0, color = "grey40", size = 8))
}

### spatial function (continuous)
spatialplot <- function(df, fillvar) {
  ggplot({{df}}) +
  geom_sf(aes(fill = {{fillvar}}), color = "white", linewidth = 0.2) +
  scale_fill_distiller(
    palette = "PuBu",
    direction = 1,) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "right")
}

### spatial function (discrete)
spatialplotCont <- function(df, fillvar) {
  ggplot({{df}}) +
  geom_sf(aes(fill = {{fillvar}}), color = "white", linewidth = 0.2) +
  scale_fill_distiller(
    palette = "PuBu",
    direction = 1,) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "right")
}

### spatial function (discrete)
spatialplotDisc <- function(df, fillvar) {
  ggplot({{df}}) +
  geom_sf(aes(fill = {{fillvar}}), color = "white", linewidth = 0.2) +
  scale_fill_brewer(
    palette = "PuBu",
    direction = 1,) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "right")
}

## spatial variables
us_map <- tigris::states(cb = TRUE, resolution = "20m", class = "sf") |>
  filter(STUSPS != "PR") |>
  tigris::shift_geometry() |>
  mutate(GEOID = str_pad(GEOID, 2, pad = "0"))
```

## EDA 1: Average Number of Firms

### By Types of Tech

```{r}
tech_2018_clean |>
  filter(!is.na(n_firms)) |>
  group_by(tech_type) |>
  summarize(avgFirms = mean(n_firms)) |>
  arrange(desc(avgFirms))
```

Among Tech Companies in the North America, `Specialized Software	` companies have the highest average number of firms at `1324.265`. `Specialized Equipment` companies report the lowest at `1189.696	` firms.

### By Tech Usage

```{r}
tech_2018_clean |>
  filter(!is.na(n_firms)) |>
  group_by(used_tech) |>
  summarize(avgFirms = mean(n_firms)) |>
  arrange(desc(avgFirms))
```

Among Tech Companies in the North America, Companies that `Did not use` their tech have the highest average number of firms at `4514.4706`. Companies that `Tested` their tech report the lowest at `144.5495` firms.

### By Types of Tech and If They Were Used

```{r}
# summary table
tech_2018_clean |>
  filter(!is.na(n_firms)) |>
  group_by(tech_type, used_tech) |>
  summarize(avgFirms = mean(n_firms)) |>
  arrange(desc(avgFirms))
```

Among Tech Companies in the North America, `Robotics` companies that `Did not use` their tech have the highest average number of firms at `5360.76368`. `Robotics` companies that `Tested` their tech report the lowest at `87.01887` firms.

```{r}
# visualization
tech_2018_clean |>
  filter(!is.na(n_firms)) |>
  group_by(tech_type, used_tech) |>
  summarize(avgFirms = mean(n_firms)) |>
  colplotCont(used_tech, avgFirms/1000, tech_type) +
  labs(title = "Average Number of Tech Firms in 2018",
       subtitle = "The vast majority of tech companies did not use their technology.",
       x = "Technology Usage",
       y = "Number of Firms (in thousands)", 
       fill = "Types of Technology", 
       caption = "Source: 2018 Annual Business Survey")
```


## EDA 2: Average Payroll per Employee

### North American Industry Classification System (naics_sector)

**Note:** `naics_sector` indices and their respective industry:

-   Total = `00`
-   Agriculture, Forestry, Fishing and Hunting = `11`
-   Mining, Quarrying, and Oil and Gas Extraction = `21`
-   Utilities = `22`
-   Construction = `23`
-   Manufacturing = `31-33`
-   Wholesale Trade = `42`
-   Retail Trade = `44-45`
-   Transportation and Warehousing = `48-49`
-   Information = `51`
-   Finance and Insurance = `52`
-   Real Estate Rental and Leasing = `53`
-   Professional, Scientific, and Technical Services = `54`
-   Management of Companies and Enterprises = `55`
-   Administrative and Support and Waste Management and Remediation Services = `56`
-   Educational Services = `61`
-   Health Care and Social Assistance = `62`
-   Arts, Entertainment, and Recreation = `71`
-   Accommodation and Food Services = `72`
-   Other Services (except Public Administration) = `81`
-   Public Administration = `92`
-   NA = `99`

**Source:** https://www.census.gov/naics/?58967?yearbck=2017

```{r}
cs_panel_clean |>
  mutate(payroll_per_emp = payroll/n_emp) |>
  group_by(naics_sector) |>
  summarize(medianPay = median(payroll_per_emp)) |>
  arrange(desc(medianPay))
```

Among industries in North America, `Utilities` report the highest median payroll per employee at `$105.31`. `Accomodation and Food Services` report the lowest at `$18.67`.

### Year

```{r}
cs_panel_clean |>
  mutate(payroll_per_emp = payroll/n_emp) |>
  group_by(year) |>
  summarize(medianPay = median(payroll_per_emp)) |>
  arrange(desc(medianPay))
```

`2021` reported the highest median payroll per employee at `$58.71`. `2017` reported the lowest at `$47.78`.

### naics_sector and Year

```{r}
# summary table
cs_panel_clean |>
  mutate(payroll_per_emp = payroll/n_emp) |>
  group_by(naics_sector, year) |>
  summarize(medianPay = median(payroll_per_emp)) |>
  arrange(desc(medianPay))
```

Among industries in North America throughout a 5-year period, `Utilities` report the highest median payroll per employee at `$111.38` during `2021`. `Accomodation and Food Services` report the lowest at `$16.18` during `2020`.

```{r}
# visualization
cs_panel_clean |>
  mutate(payroll_per_emp = payroll/n_emp) |>
  group_by(naics_sector, year) |>
  summarize(medianPay = median(payroll_per_emp)) |>
  colplotDisc(naics_sector, medianPay, year) +
  labs(title = "Median Payroll for Tech Employees",
       subtitle = "Utilities and Management have the highest pay. 2021 across all industries report the highest pay.",
       x = "Industries",
       y = "Payroll per Employee", 
       fill = "Years", 
       caption = "Source: 2018 Annual Business Survey")
```

## Spatial Viz: Revenue

```{r}
# revenue has only been reported for 2017, cleaning is necessary
revenue_2017 <- cs_panel_clean |>
  filter(year == 2017)
```

### Revenue per State

```{r}
# spatial visualization (total revenue)
state_revenue_2017 <- revenue_2017 |>
  group_by(state) |>
  summarise(total_revenue = sum(revenue/1000000000, na.rm = TRUE), .groups = "drop")

rev_map <- us_map |>
  left_join(state_revenue_2017, by = c("GEOID" = "state"))

spatialplotCont(rev_map, total_revenue) +
  labs(
    title = "Where are AI Companies Generating the Highest Revenue?",
    subtitle = "Total State Revenue in 2017",
    fill = "Total Revenue \n (in Trillions$)",
    caption = "Source: 2018 Annual Business Survey")
```

```{r}
# spatial visualization (revenue per firm)
state_revenue_2017_per_firm <- revenue_2017 |>
  group_by(state) |>
  summarise(avg_revenue = sum(revenue_per_firm/1000, na.rm = TRUE), .groups = "drop")

rev_per_firm_map <- us_map |>
  left_join(state_revenue_2017_per_firm, by = c("GEOID" = "state"))

spatialplotCont(rev_per_firm_map, avg_revenue) +
  labs(
    title = "Where are AI Companies Generating the Highest Revenue?",
    subtitle = "State Revenue per firm in 2017",
    fill = "Total Revenue \n (in Millions$)",
    caption = "Source: 2018 Annual Business Survey")
```

When looking at the maps for revenue across each states, the results vary drastically when observing it on a `total revenue` scale, and a `revenue per firm` scale. For the former, `California` proves to be the highest profitable AI state, followed by `Texas` by a difference of nearly $1 Trillion. As for the latter scale, the revenue is a lot more evenly distributed across each state, with the East Coast generating more revenue per firm compared to other regions in the US.

### Highest Profitable Industry in Each State

```{r}
# spatial visualization (highest industry revenue)
ind_revenue_2017 <- revenue_2017 |>
  group_by(state, naics_label) |>
  summarise(total_revenue = sum(revenue, na.rm = TRUE), .groups = "drop") |>
  group_by(state) |>
  slice_max(total_revenue, n = 1)

ind_rev_map <- us_map |>
  left_join(ind_revenue_2017, by = c("GEOID" = "state"))

spatialplotDisc(ind_rev_map, naics_label) +
  labs(
    title = "What Are the Highest Revenue-Generating Industries in Each State?",
    subtitle = "Highest State Revenue in 2017",
    fill = "Industry",
    caption = "Source: 2018 Annual Business Survey")
```

```{r}
# spatial visualization (highest industry revenue(per firm))
ind_per_firm_revenue_2017 <- revenue_2017 |>
  group_by(state, naics_label) |>
  summarise(avg_revenue = sum(revenue_per_firm, na.rm = TRUE), .groups = "drop") |>
  group_by(state) |>
  slice_max(avg_revenue, n = 1)

ind_per_firm_rev_map <- us_map |>
  left_join(ind_per_firm_revenue_2017, by = c("GEOID" = "state"))

spatialplotDisc(ind_per_firm_rev_map, naics_label) +
  labs(
    title = "What Are the Highest Revenue-Generating Industries in Each State?",
    subtitle = "Highest State Revenue (calculated per firm) in 2017",
    fill = "Industry",
    caption = "Source: 2018 Annual Business Survey")
```

Observing the two maps based on the same `Total Revenue` and `Revenue per Firms` scales, they paint very distinct pictures from one another. The first map shows that `Wholesale Trade` has the generate the highest amount of revenue for the vast majority of states, with `Manufacturing` following second. The second map tells a different story, `Utilities` is the most profitable industry (on a revenue per firm scale) for the vast majority of the country, with `Louisiana` and `Mississippi` differing by having `Manufacturing`, and `Hawaii` having `Finance and Insurance`.

```{r}
# summary tables
# industry revenue
ind_revenue_2017 |>
  arrange(desc(total_revenue))

# industry revenue per firm
ind_per_firm_revenue_2017 |>
  arrange(desc(avg_revenue))
```

Interestingly enough, state `11`, or the `District of Columbia` has the most profitable industry per firm, despite it not being `Utilities` like all of the other states.
