---
title: "David's EDA"
format: 
  html:
    toc: true
    code-fold: true
---

## Motivations

In this document, I will be looking at the evolution of AI. Since we only have data for 2018 on AI-usage, I would look within those industries leading the AI race, how much have their payroll evolve. We excluded the *total* observations, as well as the *unknown* category.

## Introduction

```{r, setup, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(plotly)
library(dplyr)
library(tigris)
library(shiny)
library(leaflet)
library(sf)
library(stringr)

source(here::here("data", "data_cleaning.R"))

## Theme Settings
theme_set(theme_minimal(base_size = 12))

scale_use_level <- function(...) {
  scale_color_brewer(palette = "Dark2", name = NULL, ...)
}

scale_use_level_fill <- function(...) {
  scale_fill_brewer(palette = "Dark2", name = NULL, ...)
}
```


## Who's the top industry using AI?

```{r}
tech_2018_clean |> 
  filter(tech_type == "artificial intelligence") |> 
  group_by(naics_sector) |> 
  ggplot(aes(x = used_tech , fill =use_level )) +
  geom_density(alpha=0.5) +
  scale_use_level_fill() +
  theme_set()

tech_2018_clean |> 
  filter(tech_type == "artificial intelligence") |> 
  group_by(state, naics_sector) |> 
  ggplot(aes(x = used_tech , fill  = naics_label)) +
  geom_density(alpha=0.5) +
  scale_use_level_fill() +
  theme_set()
```




> Based on the payroll/n_emp and ai usage, how much is the revenue changing? Are those firms who are paying their employee more seeing more gains? Or is it a matter of company size?


## The AI Wage Premium

Does the AI wage premium holds constant as firms grow in size?

```{r}
## wage premium story
tech_2018_clean |>
  filter(tech_type == "artificial intelligence",
         used_tech != "Unknown",
         naics_sector != "Total", 
         payroll > 0,
         n_emp > 0) |>
  mutate(
    ai_used = if_else(used_tech == "Used", "AI Adopters", "Non-Adopters"),
    payroll_per_emp = payroll / n_emp) |> 
  ggplot(aes(x = n_emp, y = payroll_per_emp, color = ai_used)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(se = FALSE, linewidth = 1.5) + 
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(
    labels = scales::label_dollar(suffix = "K"), 
    breaks = seq(0, 300, by = 50)
  ) +  
  scale_use_level() +
  labs(
    title = "Does AI Pay?",
    subtitle = "Evidence of higher payroll standards among technology adopters across firm scales",
    x = "Number of Employees (Log Scale)",
    y = "Avg. Payroll per Employee (USD)",
    color = "Adoption Status",
    caption = "Source: 2018 Annual Business Survey (ABS) | Author: David Rios\nNote: Payroll standardized to thousands of dollars per employee. Trend lines calculated via local regression (LOESS)."
  ) +
  theme_minimal() +
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8))
```
## Patterns within Industries


### Company Size Panel Clean (2017-2021)

> While adoption alone signals a premium, intensity is the true driver. In sectors like Finance and Information, high implementation leads to the most significant wage increases.

```{r explo}
# Management sector
tech_2018_clean |> 
  filter(naics_sector == "Management", tech_type == "artificial intelligence") |> 
  group_by(use_level) |> 
  summarise(firm_count = sum(n_firms, na.rm = TRUE)) |> 
  arrange(use_level)

```

## Patterns within Industries

```{r industry_patterns, message=FALSE, warning=FALSE}
# Ensure the scaling matches your data source
# If payroll is $80,000, scale = 0.001 makes it $80K
# If payroll is already 80, scale = 0.001 makes it $0.08K (incorrect)

tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    naics_sector %in% c("Information", "Finance and Insurance", 
                        "Professional Services", "Management"),
    use_level != "Don't know",
    n_emp > 0, 
    payroll > 0
  ) |>
  mutate(
    payroll_per_emp = payroll / n_emp,
    use_level = fct_relevel(use_level,
                            "Did not use",
                            "Tested, but did not use in production or service",
                            "Low use", "Moderate use", "High use")
  ) |>
  group_by(naics_sector, use_level) |> 
  summarise(
    wpay = weighted.mean(payroll_per_emp, w = n_emp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = use_level, y = wpay, fill = use_level)) +
  geom_col(width = 0.7) +
  scale_use_level_fill() +
  scale_x_discrete(labels = c(
    "Did not use" = "None",
    "Tested, but did not use in production or service" = "Tested",
    "Low use" = "Low", "Moderate use" = "Moderate", "High use" = "High"
  )) +
  # This line is the most common point of failure:
  scale_y_continuous(labels = scales::label_dollar( suffix = "K")) +
  facet_wrap(~naics_sector, scales = "free_y") + 
  labs(
    title = "AI Intensity and Payroll",
    subtitle = "Higher levels of AI implementation correlate with significant wage premiums",
    x = NULL, 
    y = "Avg. Payroll per Employee",
    caption = "Source: 2018 ABS. Note: 'High Use' category for Management contains zero or suppressed reporting firms."
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 8),
    strip.text = element_text(face = "bold")
  )
```


## AI Adoption Spatial

> Note: industry sectors with more than 10 missing states were excluded from this analysis. These sectors are Agriculture, Mining, and Utilities. The remaining sectors provide a more reliable look at AI adoption trends. AI adoption rates are rendered to two decimal places to capture variations in sectors where adoption is emerging but still below 1%.

## AI Adoption Spatial

```{r maps}
# --- PREPARATION: State Lookups ---
fips_to_name <- tibble(
  state_fips = c("01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", "38", "39", "40", "41", "42", "44", "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", "56"),
  state_name = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming")
)

# --- STEP 1: Geometry (Silent & Clean) ---
sf_use_s2(FALSE)
states_sf <- suppressMessages(suppressWarnings(
  st_as_sf(maps::map("state", plot = FALSE, fill = TRUE)) |>
    mutate(state_name = str_to_title(word(ID, 1, sep = ":"))) |>
    st_make_valid() |> 
    group_by(state_name) |> 
    summarise(geometry = st_union(geom), .groups = "drop") |>
    st_transform(4326)
))

# --- STEP 2: AI Rates (Summarized & Defensive) ---
ai_rates <- tech_2018_clean |>
  ungroup() |> 
  filter(tech_type == "artificial intelligence") |>
  mutate(naics_sector = as.character(naics_sector)) |>
  group_by(state, naics_sector) |>
  summarise(
    ai_adoption_rate = sum(n_firms[used_tech == "Used"], na.rm = TRUE) / 
      sum(n_firms, na.rm = TRUE),
    .groups = "drop"
  )

# --- STEP 3: Payroll (Matching by Cleaned Sector Names) ---
payroll_stats <- cs_panel_clean |>
  ungroup() |> 
  group_by(state, naics_sector) |>
  summarise(
    mean_payroll_per_emp = mean(payroll_per_emp, na.rm = TRUE), # Raw USD
    .groups = "drop"
  )

# --- STEP 4: Final Merge & Industry Filter ---
state_ai_spatial <- ai_rates |>
  left_join(payroll_stats, by = c("state", "naics_sector")) |>
  left_join(fips_to_name, by = c("state" = "state_fips")) |>
  filter(!is.na(state_name))

# Identify industries with enough data to show in the dropdown
valid_industries <- state_ai_spatial |>
  group_by(naics_sector) |>
  summarise(missing = sum(is.na(ai_adoption_rate)), .groups = "drop") |>
  filter(missing <= 10, !naics_sector %in% c("Unknown", "Total")) |>
  pull(naics_sector)

# --- STEP 5: Shiny App ---
ui <- fluidPage(
  titlePanel("US AI Adoption & Payroll by Sector (2018)"),
  sidebarLayout(
    sidebarPanel(
      selectInput("industry_choice", "Select Industry Sector:", 
                  choices = sort(valid_industries),
                  selected = "Professional Services"),
      hr(),
      helpText("Hover over states for metrics. Grey indicates insufficient data.")
    ),
    mainPanel(leafletOutput("map", height = "750px"))
  )
)

server <- function(input, output, session) {
  
  filtered_data <- reactive({
    req(input$industry_choice)
    state_ai_spatial |>
      filter(naics_sector == input$industry_choice) |>
      left_join(states_sf, by = "state_name") |>
      st_as_sf() |>
      filter(!st_is_empty(geometry))
  })
  
  output$map <- renderLeaflet({
    leaflet() |>
      addProviderTiles(providers$CartoDB.Positron) |>
      setView(lng = -98.58, lat = 39.82, zoom = 4)
  })
  
  observe({
    data <- filtered_data()
    req(nrow(data) > 0)
    
    pal <- colorNumeric("YlOrRd", 
                        domain = c(0, max(data$ai_adoption_rate * 100, na.rm = TRUE)), 
                        na.color = "#D3D3D3")
    
    labels <- sprintf(
      "<strong>%s</strong><br/>
       Sector: %s<br/>
       AI Adoption: <b>%s</b><br/>
       Payroll/Emp: <b>%s</b>",
      data$state_name, 
      data$naics_sector, 
      ifelse(is.na(data$ai_adoption_rate), 
             "Insufficient Data", 
             paste0(round(data$ai_adoption_rate * 100, 2), "%")),
      # Applied / 1000 fix here for raw USD to 'k' display
      ifelse(is.na(data$mean_payroll_per_emp), 
             "N/A", 
             paste0("$", formatC(data$mean_payroll_per_emp / 1000, format="f", digits=1, big.mark=","), "k"))
    ) |> lapply(htmltools::HTML)
    
    leafletProxy("map", data = data) |>
      clearShapes() |>
      addPolygons(
        fillColor = ~pal(ai_adoption_rate * 100),
        fillOpacity = 0.8, color = "white", weight = 1,
        highlightOptions = highlightOptions(weight = 3, color = "#666", bringToFront = TRUE),
        label = labels
      ) |>
      clearControls() |>
      addLegend("bottomright", pal = pal, values = ~ai_adoption_rate * 100,
                title = "AI Adoption (%)", labFormat = labelFormat(suffix = "%"),
                na.label = "No Data")
  })
}

shinyApp(ui, server)
```

