---
title: "David's EDA"
format: html
---


## Introduction

```{r, package, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(plotly)
library(dplyr)
library(tigris)
library(shiny)
library(leaflet)
library(sf)
library(stringr)

source(here::here("data", "data_cleaning.R"))
```

```{r cleaning}
tech_2018_clean <- tech_2018_clean |> 
  mutate(
    naics_sector = fct_recode(as.factor(naics_sector),
                              "Total"                          = "00",
                              "Agriculture"       = "11",
                              "Mining"                         = "21",
                              "Utilities"                      = "22",
                              "Construction"                   = "23",
                              "Manufacturing"                  = "31-33",  
                              "Wholesale Trade"                = "42",
                              "Retail Trade"                   = "44-45",  
                              "Transportation" = "48-49",  
                              "Information"                    = "51",
                              "Finance and Insurance"          = "52",
                              "Real Estate"                    = "53",
                              "Professional Services"          = "54",
                              "Management"                     = "55",
                              "Administrative Services"        = "56",
                              "Educational Services"           = "61",
                              "Health Care"                    = "62",
                              "Arts and Recreation"            = "71",
                              "Food Services"= "72",
                              "Other Services"                 = "81",
                              "Unknown"                        = "99"     
    )
  )


tech_2018_clean |> 
  group_by(naics_sector) |> 
  count()

## setting theme for plots
theme_set(theme_minimal(base_size = 12))

# reusable color scale for use levels
scale_use_level <- function(...) {
  scale_color_brewer(palette = "Dark2", name = NULL, ...)
}
scale_use_level_fill <- function(...) {
  scale_fill_brewer(palette = "Dark2", name = NULL, ...)
}
```

## Descriptive Statistics

```{r prueba}
merged_panel %>%
  summarize(
    missing_ai_rate = sum(is.na(ai_adoption_rate)),
    nan_ai_rate = sum(is.nan(ai_adoption_rate)),
    zero_ai_total = sum(ai_firms_total == 0, na.rm = TRUE))

merged_panel <- merged_panel %>%
  mutate(
    ai_adoption_rate = ifelse(ai_firms_total == 0, NA, ai_adoption_rate))
```



> Based on the payroll/n_emp and ai usage, how much is the revenue changing? Are those firms who are paying their employee more seeing more gains? Or is it a matter of company size?




## Tech 2018 Data Set: ABS

### Size vs. AI

```{r}
# 1. What tech types and use levels do we have?
tech_2018_clean |> count(tech_type, used_tech)

# 2. Revenue is all zeros — confirm and check payroll coverage
tech_2018_clean |>
  summarise(
    pct_zero_revenue = mean(revenue == 0) * 100,
    pct_zero_payroll = mean(payroll == 0) * 100,
    median_payroll   = median(payroll[payroll > 0], na.rm = TRUE)
  )

# 3. AI-specific: payroll by usage level
tech_2018_clean |>
  filter(tech_type == "artificial intelligence",
         used_tech != "Unknown",
         naics_sector != "00",
         payroll > 0) |>
  group_by(used_tech, naics_sector) |>
  summarise(
    total_firms   = sum(n_firms,  na.rm = TRUE),
    total_emp     = sum(n_emp,    na.rm = TRUE),
    total_payroll = sum(payroll,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(payroll_per_emp = total_payroll / total_emp) |>
  arrange(used_tech, desc(payroll_per_emp))

# Check if payroll is at firm or employee level
tech_2018_clean |>
  filter(tech_type == "artificial intelligence",
         naics_sector == "00",   # total economy
         used_tech == "Used") |>
  select(state, n_firms, n_emp, payroll) |>
  mutate(payroll_per_emp = payroll / n_emp) |>
  arrange(desc(payroll_per_emp))



## wage premium story
tech_2018_clean |>
  select(-revenue) |> 
  filter(tech_type    == "artificial intelligence",
    used_tech    != "Unknown",
    naics_sector != "00",
    naics_sector != "99",
    payroll > 0,
    n_emp   > 0) |>
  mutate(
    ai_used         = if_else(used_tech == "Used", "AI Adopters", "Non-Adopters"),
    payroll_per_emp = payroll / n_emp) |>
  ggplot(aes(x = n_emp, y = payroll_per_emp, color = ai_used)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_x_log10(labels = scales::comma) +
scale_use_level () +
  scale_use_level_fill() +  
  labs(
    title    = "Firm Size, AI Adoption, and Payroll",
    subtitle = "Each point is a state × industry cell",
    x        = "Number of Employees (log scale)",
    y        = "Payroll per Employee, ($K)",
    caption  = "Source: 2018 Annual Business Survey") +
  theme_minimal() +
  theme(legend.position = "top")

```

## Company Size Panel Clean (2017-2021)

```{r}
tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    # FIX: Filter by the NEW names, not the old codes
    naics_sector %in% c("Information", "Finance and Insurance", "Professional Services", "Management"),
    use_level != "Don't know",
    n_emp > 0, 
    payroll > 0
  ) |>
  mutate(
    payroll_per_emp = payroll / n_emp,
    use_level = fct_relevel(use_level,
      "Did not use",
      "Tested, but did not use in production or service",
      "Low use", "Moderate use", "High use")
  ) |>
  group_by(naics_sector, use_level) |> # Use naics_sector directly
  summarise(
    wpay = weighted.mean(payroll_per_emp, w = n_emp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = use_level, y = wpay, fill = use_level)) +
  geom_col(width = 0.7) +
  scale_fill_brewer(palette = "Dark2", name = NULL) +
  scale_x_discrete(labels = c(
    "Did not use" = "None",
    "Tested, but did not use in production or service" = "Tested",
    "Low use" = "Low", "Moderate use" = "Mod", "High use" = "High"
  )) +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~naics_sector, scales = "free_y") + # Facet by the sector name
  labs(
    title    = "AI Intensity and Payroll: Pattern Holds Within Industries",
    subtitle = "Weighted mean payroll per employee ($000s) by AI use level, 2018",
    x = NULL, y = "Payroll per Employee ($000s)",
    caption  = "Source: 2018 ABS. Weighted by employment."
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 7))

```

## AI Adoption Spatial

> Note: industry sectors with more than 10 missing states were excluded from this analysis. These sectors are Agriculture, Mining, and Utilities. The remaining sectors provide a more reliable look at AI adoption trends. AI adoption rates are rendered to two decimal places to capture variations in sectors where adoption is emerging but still below 1%.

## AI Adoption Spatial

```{r maps}
fips_to_name <- tibble(
  state_fips = c("01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", "38", "39", "40", "41", "42", "44", "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", "56"),
  state_name = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming")
)

# --- STEP 1: Geometry (Silent & Clean) ---
sf_use_s2(FALSE)
states_sf <- suppressMessages(suppressWarnings(
  st_as_sf(maps::map("state", plot = FALSE, fill = TRUE)) |>
    mutate(state_name = str_to_title(word(ID, 1, sep = ":"))) |>
    st_make_valid() |> 
    group_by(state_name) |> 
    summarise(geometry = st_union(geom), .groups = "drop") |>
    st_transform(4326)
))

# --- STEP 2: AI Rates (Summarized & Defensive) ---
ai_rates <- tech_2018_clean |>
  ungroup() |> 
  filter(tech_type == "artificial intelligence") |>
  mutate(naics_sector = as.character(naics_sector)) |>
  group_by(state, naics_sector) |>
  summarise(
    ai_adoption_rate = sum(n_firms[used_tech == "Used"], na.rm = TRUE) / 
                       sum(n_firms, na.rm = TRUE),
    .groups = "drop"
  )

# --- STEP 3: Payroll (Standardizing Sector Names) ---
payroll_stats <- cs_panel_clean |>
  ungroup() |> 
  mutate(sector_name = case_when(
    naics_sector == "11" ~ "Agriculture",
    naics_sector == "21" ~ "Mining",
    naics_sector == "22" ~ "Utilities",
    naics_sector == "23" ~ "Construction",
    naics_sector == "31-33" ~ "Manufacturing",
    naics_sector == "42" ~ "Wholesale Trade",
    naics_sector == "44-45" ~ "Retail Trade",
    naics_sector == "48-49" ~ "Transportation",
    naics_sector == "51" ~ "Information",
    naics_sector == "52" ~ "Finance and Insurance",
    naics_sector == "53" ~ "Real Estate",
    naics_sector == "54" ~ "Professional Services",
    naics_sector == "55" ~ "Management",
    naics_sector == "56" ~ "Administrative Services",
    naics_sector == "61" ~ "Educational Services",
    naics_sector == "62" ~ "Health Care",
    naics_sector == "71" ~ "Arts and Recreation",
    naics_sector == "72" ~ "Food Services",
    naics_sector == "81" ~ "Other Services",
    TRUE ~ "Unknown"
  )) |>
  group_by(state, sector_name) |>
  summarise(
    mean_payroll_per_emp = mean(payroll_per_emp / 1000, na.rm = TRUE),
    .groups = "drop")

# --- STEP 4: Final Merge & Industry Filter ---
state_ai_spatial <- ai_rates |>
  left_join(payroll_stats, by = c("state" = "state", "naics_sector" = "sector_name")) |>
  left_join(fips_to_name, by = c("state" = "state_fips")) |>
  filter(!is.na(state_name))

# Data Quality Check: Filter out industries with > 10 missing states
valid_industries <- state_ai_spatial |>
  group_by(naics_sector) |>
  summarise(missing = sum(is.na(ai_adoption_rate)), .groups = "drop") |>
  filter(missing <= 10, !naics_sector %in% c("Unknown", "Total")) |>
  pull(naics_sector)

# --- STEP 5: Shiny App ---
ui <- fluidPage(
  titlePanel("US AI Adoption & Payroll by Sector (2018)"),
  sidebarLayout(
    sidebarPanel(
      selectInput("industry_choice", "Select Industry Sector:", 
                  choices = sort(valid_industries),
                  selected = "Professional Services"),
      hr(),
      helpText("Hover over states for metrics. Grey indicates insufficient data.")
    ),
    mainPanel(leafletOutput("map", height = "750px"))
  )
)

server <- function(input, output, session) {
  
  filtered_data <- reactive({
    req(input$industry_choice)
    state_ai_spatial |>
      filter(naics_sector == input$industry_choice) |>
      left_join(states_sf, by = "state_name") |>
      st_as_sf() |>
      filter(!st_is_empty(geometry))
  })

  output$map <- renderLeaflet({
    leaflet() |>
      addProviderTiles(providers$CartoDB.Positron) |>
      setView(lng = -98.58, lat = 39.82, zoom = 4)
  })

  observe({
    data <- filtered_data()
    req(nrow(data) > 0)
    
    # Custom palette that handles NAs as grey
    pal <- colorNumeric("YlOrRd", 
                        domain = c(0, max(data$ai_adoption_rate * 100, na.rm = TRUE)), 
                        na.color = "#D3D3D3")
    
    labels <- sprintf(
      "<strong>%s</strong><br/>
       Sector: %s<br/>
       AI Adoption: <b>%s</b><br/>
       Payroll/Emp: <b>%s</b>",
      data$state_name, data$naics_sector, 
      ifelse(is.na(data$ai_adoption_rate), "Insufficient Data", paste0(round(data$ai_adoption_rate * 100, 2), "%")),
      ifelse(is.na(data$mean_payroll_per_emp), "N/A", paste0("$", formatC(data$mean_payroll_per_emp, format="f", digits=1, big.mark=","), "k"))
    ) |> lapply(htmltools::HTML)

    leafletProxy("map", data = data) |>
      clearShapes() |>
      addPolygons(
        fillColor = ~pal(ai_adoption_rate * 100),
        fillOpacity = 0.8, color = "white", weight = 1,
        highlightOptions = highlightOptions(weight = 3, color = "#666", bringToFront = TRUE),
        label = labels
      ) |>
      clearControls() |>
      addLegend("bottomright", pal = pal, values = ~ai_adoption_rate * 100,
                title = "AI Adoption (%)", labFormat = labelFormat(suffix = "%"),
                na.label = "No Data")
  })
}

shinyApp(ui, server)
```

