---
title: "David's Exploration"
format: 
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Motivations

In this document, I will be looking at the evolution of AI. Since we only have data for 2018 on AI-usage, I would look within those industries leading the AI race, how much have their payroll evolve. We excluded the *total* observations, as well as the *unknown* category.

## Introduction

```{r setup, message=FALSE}
library(here)
library(tidyverse)
library(tigris)
library(stringr)
library(scales)
library(ggrepel)
library(tidymodels)
library(ranger)
library(patchwork)
library(leaflet) 
library(forcats)
library(rpart)
library(tibble)   
library(rpart.plot)
library(paletteer)

source(here::here("data", "data_cleaning.R"))

## Theme Settings
theme_set(
  theme_minimal(base_size = 12) + 
    theme(
      legend.position = "right",
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30"),
      plot.caption = element_text(hjust = 0, color = "grey40", size = 10),
      panel.grid.minor = element_blank()))
```


## Who's the top industry using AI?

I am interested in seeing who was the top industry in the "AI race" in 2018.

```{r, top_ai}
ai_industry_leaders <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence", 
    !is.na(used_tech), 
    used_tech != "Unknown") |>
  group_by(naics_label) |>
  summarise(
    total_firms = sum(n_firms, na.rm = TRUE),
    ai_firms = sum(n_firms[used_tech == "Used"], na.rm = TRUE),
    .groups = "drop") |>
  mutate(ai_rate = ai_firms / total_firms) |>
  filter(total_firms > 0) |>
  arrange(desc(ai_rate))

ai_industry_leaders |> 
  filter(ai_firms>0) |> 
  ggplot(aes(x = reorder(naics_label, ai_rate), y = ai_rate, fill = ai_rate)) +
  geom_col() + 
  scale_fill_distiller(
    palette = "PuBu",
    direction = 1)+  coord_flip() + 
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Which Industries Led the AI Race in 2018?",
    subtitle = "Percentage of firms reporting AI Use",
    x = NULL,
    y = "AI Adoption Rate",
    caption = "Management took the lead in 2018, with an AI implementation of 7.7%. \nSource: 2018 Annual Business Survey (ABS) | Author: David Rios ") +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(), 
        axis.ticks.y = element_blank(),
        legend.position = "")
```

## Payroll Evolution

### Level Change

I then got interested in analyzing what's the payroll evolution by industry, but this was not telling me much so I also looked at the Payroll Percentage Change instead, and I took 2017 as the base year.

```{r payroll_lvl}
tech_2018_summary <- tech_2018_clean |> 
  group_by(year, used_tech, use_level) |> 
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop")

payroll_all <- cs_panel_clean |>
  group_by(year, naics_label) |>
  summarise(total_payroll = sum(payroll * 1000, na.rm = TRUE), .groups = "drop")

top_4_industries <- ai_industry_leaders |> slice_max(ai_rate, n = 4) |> pull(naics_label)
payroll_top4 <- payroll_all |> filter(naics_label %in% top_4_industries)

ggplot() +
  geom_line(data = payroll_all,  
            aes(x = year, y = total_payroll, group = naics_label), 
            color = "grey85", linewidth = 0.5, alpha = 0.6) +
  geom_line(data = payroll_top4, 
            aes(x = year, y = total_payroll, color = naics_label, group = naics_label), 
            linewidth = 1) +
  geom_point(data = payroll_top4, 
             aes(x = year, y = total_payroll, color = naics_label), 
             size = 1.5) +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-9, suffix = " B", accuracy = 1)) +
  scale_color_viridis_d(option = "H") +
  labs(
    title = "Payroll Evolution",
    subtitle = "Highlighting the 4 industries with the highest AI adoption rates in 2018",
    x = "Year",
    y = "Total Annual Payroll (Billions)",
    color = NULL,
    caption = "Healthcare is the top industry by total payroll, despite showing low AI adoption rates in 2018.  \nSource: 2018 Annual Business Survey (ABS) | Author: David Rios "
  ) + theme(legend.position = "top")
```

### Percentage Change

```{r payroll_pct}
## indexed payroll growth
payroll_growth <- payroll_all |>
  arrange(naics_label, year) |>
  group_by(naics_label) |>
  mutate(
    payroll_growth_pct = total_payroll / first(total_payroll)) |> #base year: 2017
  ungroup() |>
  mutate(naics_label = str_wrap(naics_label, width = 35))


## Top AI industries
top_4_industries <- ai_industry_leaders |>
  slice_max(ai_rate, n = 4) |>
  pull(naics_label)

payroll_growth_top4 <- payroll_growth |>
  filter(naics_label %in% top_4_industries)


## Bottom AI industries
w_4_industries <- ai_industry_leaders |>
  slice_min(ai_rate, n = 4) |>
  pull(naics_label)

payroll_growth_w4 <- payroll_growth |>
  filter(naics_label %in% w_4_industries)


# plots
p1_top4_ai <- ggplot() +
  geom_line(
    data = payroll_growth,
    aes(x = year, y = payroll_growth_pct, group = naics_label),
    color = "grey85", linewidth = 0.5, alpha = 0.6) +
  geom_line(
    data = payroll_growth_top4,
    aes(x = year, y = payroll_growth_pct, color = naics_label),
    linewidth = 1) +
  geom_text_repel(
    data = payroll_growth_top4 |> filter(year == max(year)),
    aes(x = year, y = payroll_growth_pct, label = naics_label, color = naics_label),
    nudge_x = 1,
    show.legend = FALSE,
    size = 3) +
  scale_y_continuous(labels = percent_format()) +
    scale_color_viridis_d(option = "D") +
  labs(
    title = "AI-Leading Industries",
    subtitle = "Indexed to 2017 total payroll",
    x = "Year",
    y = "Payroll (2017 = 1)") +
  theme_minimal() +
 theme(
    legend.position = "none")

p2_worse_ai <- ggplot() +
  geom_line(
    data = payroll_growth,
    aes(x = year, y = payroll_growth_pct, group = naics_label),
    color = "grey85", linewidth = 0.5, alpha = 0.6) +
  geom_line(
    data = payroll_growth_w4,
    aes(x = year, y = payroll_growth_pct, color = naics_label),
    linewidth = 1) +
  geom_point(
    data = payroll_growth_w4,
    aes(x = year, y = payroll_growth_pct, color = naics_label),
    size = 1) +
  geom_text_repel(
    data = payroll_growth_w4 |> filter(year == max(year)),
    aes(
      x = year,
      y = payroll_growth_pct,
      label = naics_label,
      color = naics_label),
    nudge_x = 1,
    show.legend = FALSE,
    size = 3) +
  scale_y_continuous(labels = percent_format()) +
    scale_color_viridis_d(option = "E") +
  labs(
    title = "Lower AI-Adoption Industries",
    subtitle = "Indexed to 2017 total payroll",
    x = "Year",
    y = NULL,
    caption = "Source: 2018 ABS + CS Panel Data") +
  theme_minimal() +
  theme(legend.position = "none")

## final plot combining both plots above 
final_dis <- p1_top4_ai + p2_worse_ai +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = "Payroll Growth Comparison",
    theme = theme(
      plot.title = element_text(face = "bold", size = 14)))

final_dis
```


## Revenue

We only have revenue for the yeaer 2017, so I also looked at the revenue for that year by industry. 

::: {.callout-note}
**FACT CHECK:** The [US Census Bureau](https://www.census.gov/library/stories/2020/11/2017-economic-census-data-breaks-out-industry-growth-changes-by-state.html) also reported this exact figure as the one in my visualization.
:::

```{r revenue_ai}
revenue_2017 <- cs_panel_clean |>
  filter(year == 2017) |>
  group_by(naics_label) |>
  summarise(total_revenue = sum(revenue * 1000, na.rm = TRUE), .groups = "drop") |>
  filter(total_revenue > 0) |>
  arrange(desc(total_revenue))

#  2017 snapshot (in Trillions)
ggplot(revenue_2017, aes(x = total_revenue, fill = total_revenue, y = reorder(naics_label, total_revenue))) +
  geom_col(width = 0.9) + 
  scale_x_continuous(
    labels = scales::label_dollar(scale = 1e-12, suffix = "T", accuracy = 1),
    breaks = seq(0, 10e12, by = 1e12),
    expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Total Industry Revenue in 2017",
    subtitle = "A snapshot of financial volume right before the 2018 AI Survey",
    x = "Total Revenue (Trillions)",
    y = NULL,
    caption = "Source: 2018 Annual Business Survey (ABS) | Author: David Rios ") +  theme_minimal(base_size = 12) + 
  scale_fill_distiller(
    palette = "PuBu",
    direction = 1) + 
  theme(
    legend.position = "",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    plot.caption = element_text(hjust = 0, color = "grey40", size = 8),
    panel.grid.minor = element_blank())
```


## The AI Wage Premium

The evidence suggests that AI-adopting firms consistently pay higher average wages per employee, but the gap shrinks as firms become larger, indicating diminishing differences in wage premia at scale.

Takeaways:

  - Smaller firms show a clearer wage advantage for AI adopters
  - Larger firms exhibit a weaker AI–wage gap
  - Overall pattern suggests heterogeneous returns to AI adoption across firm size

```{r, wage_premium}

## wage premium story
tech_2018_clean |>
  filter(tech_type == "artificial intelligence",
         used_tech != "Unknown",
         naics_sector != "Total", 
         payroll > 0,
         n_emp > 0) |>
  mutate(
    ai_used = if_else(used_tech == "Used", "AI Adopters", "Non-Adopters"),
    payroll_per_emp = payroll / n_emp) |> 
  ggplot(aes(x = n_emp, y = payroll_per_emp, color = ai_used)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(se = FALSE, linewidth = 1.5) +  # method = "mgcv",
  scale_x_log10(labels = scales::comma) +
  scale_y_continuous(
    labels = scales::label_dollar(suffix = "K"), 
    breaks = seq(0, 300, by = 50)) +  
    scale_color_viridis_d(option = "E") +
  labs(
    title = "Does AI Pay?",
    subtitle = "Evidence of higher payroll standards among technology adopters across firm scales",
    x = "Number of Employees (Log Scale)",
    y = "Avg. Payroll per Employee (USD)",
    color = "Adoption Status",
    caption = "Payroll standardized to thousands of dollars per employee. Trend lines calculated via local regression (LOESS). \n Source: 2018 Annual Business Survey (ABS) | Author: David Rios"
  ) +
  theme_minimal() + 
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8))
```


## AI Hubs in the US

Geographically, where is AI growing the most rapidly? Or where is it AI adoption rate higher?

```{r ai_maps}
# AI Adoption Rate by State
state_ai_adoption <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    !is.na(used_tech),
    used_tech != "Unknown") |>
  group_by(state) |>
  summarise(
    total_firms = sum(n_firms, na.rm = TRUE),
    ai_firms = sum(n_firms[used_tech == "Used"], na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    state = str_pad(state, 2, pad = "0"),
    ai_rate = if_else(total_firms > 0, ai_firms / total_firms, NA_real_)) |>
  filter(total_firms > 0)

# Fetch map boundaries and shift AK/HI to the bottom
us_map <- tigris::states(cb = TRUE, resolution = "20m", class = "sf") |>
  filter(STUSPS != "PR") |>
  tigris::shift_geometry() |>
  mutate(GEOID = str_pad(GEOID, 2, pad = "0"))

map_data <- us_map |>
  left_join(state_ai_adoption, by = c("GEOID" = "state"))


ggplot(map_data) +
  geom_sf(aes(fill = ai_rate), color = "white", linewidth = 0.2) +
  scale_fill_distiller(
    palette = "PuBu",
    direction = 1,
    labels = scales::percent_format(accuracy = 1),
    na.value = "grey95") +
  labs(
    title = "Where are the AI Hubs?",
    subtitle = "AI Adoption Rate by State in 2018",
    fill = "Adoption\nRate",
    caption = "Source: 2018 Annual Business Survey (ABS)") +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "right")
```

## Wage Gaps and Premium

But is it worth to pay the cost? Are the revenues generated from AI higher than the cost of it?

```{r, data_prep_cs}
tech_2018_summary <- tech_2018_clean |>
  filter(
    !is.na(used_tech),
    MOTUSETECH_LABEL == "All firms",
    IMPACTWF_U_LABEL == "All firms"    ) |>
  group_by(year, tech_type, used_tech, use_level) |>
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop") |>
  mutate(
    emp_per_firm = if_else(total_firms > 0, total_emp / total_firms, NA_real_),
    avg_wage     = if_else(total_emp > 0, total_payroll / total_emp, NA_real_))

industry_tech_summary <- tech_2018_clean |> 
  filter(
    !is.na(used_tech), 
    MOTUSETECH_LABEL == "All firms",   
    IMPACTWF_U_LABEL == "All firms"   ) |> 
  group_by(year, naics_label, tech_type, used_tech, use_level) |> 
  summarise(
    total_firms   = sum(n_firms, na.rm = TRUE),
    total_emp     = sum(n_emp, na.rm = TRUE),
    total_payroll = sum(payroll, na.rm = TRUE),
    .groups       = "drop") |> 
  mutate(
    emp_per_firm = if_else(total_firms > 0, total_emp / total_firms, NA_real_),
    avg_wage     = if_else(total_emp > 0, total_payroll / total_emp, NA_real_))

ai_industry_wages <- industry_tech_summary |>
  filter(
    tech_type == "artificial intelligence",
    used_tech %in% c("Did not use", "Used")) |>
  group_by(naics_label, used_tech) |>
  summarise(
    total_emp = sum(total_emp, na.rm = TRUE),
    total_payroll = sum(total_payroll, na.rm = TRUE),
    .groups = "drop") |>
  mutate(avg_wage = if_else(total_emp > 0, total_payroll / total_emp, NA_real_)) |>
  filter(total_emp > 0) 

# one row per industry with both groups side-by-side
wage_wide <- ai_industry_wages |>
  filter(used_tech %in% c("Did not use", "Used")) |>
  group_by(naics_label) |>
  filter(n() == 2) |>   # only industries where BOTH groups exist
  ungroup() |>
  pivot_wider(
    id_cols    = naics_label,
    names_from = used_tech,
    values_from = c(avg_wage, total_emp, total_payroll)) |>
  rename(
    wage_no_ai = `avg_wage_Did not use`,
    wage_ai    = `avg_wage_Used`,
    emp_no_ai  = `total_emp_Did not use`,
    emp_ai     = `total_emp_Used`) 

 wage_wide <- wage_wide |>
  mutate(
    wage_gap = wage_ai - wage_no_ai,
    premium_pct = if_else(wage_no_ai > 0, wage_gap / wage_no_ai, NA_real_),
    total_emp = emp_no_ai + emp_ai,
    ai_emp_share = if_else(total_emp > 0, emp_ai / total_emp, NA_real_))
```


```{r, plots_cs}
##  The AI Wage Gap by Industry
dumbbell_long <- wage_wide |>
  pivot_longer(
    cols = c(wage_no_ai, wage_ai),
    names_to = "group",
    values_to = "avg_wage") |>
  mutate(
    group = if_else(group == "wage_ai", "AI Adopters", "Non-Adopters"),
    naics_label = reorder(naics_label, wage_gap))

ggplot() +
  geom_segment(
    data = wage_wide |> mutate(naics_label = reorder(naics_label, wage_gap)),
    aes(x = wage_no_ai, xend = wage_ai,
        y = naics_label, yend = naics_label),
    color = "grey75", linewidth = 1.3) +
  geom_point(
    data = dumbbell_long,
    aes(x = avg_wage, y = naics_label, color = group),
    size = 4) +
  scale_color_viridis_d(option = "D") +
  scale_x_continuous(labels = scales::label_dollar(suffix = "K")) +
  labs(
    title = "The AI Wage Gap by Industry",
    subtitle = "AI adopters vs. non-adopters (2018)",
    x = "Avg. Annual Payroll per Employee (Thousands USD)",
    y = NULL,
    color = NULL,
    caption = "Source: 2018 ABS") +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(color = "grey30"),
        plot.caption = element_text(hjust = 0, color = "grey40", size = 8))



## AI Wage Premium: Who Benefits Most?

wage_wide |>
  mutate(
    naics_label  = reorder(naics_label, premium_pct),
    premium_sign = if_else(premium_pct >= 0, "Positive", "Negative")
  ) |>
  ggplot(aes(x = premium_pct, y = naics_label, fill = premium_sign)) +
  geom_col(show.legend = FALSE, width = 0.75) +
  geom_vline(xintercept = 0, color = "grey40", linewidth = 0.5) +
  # Label the bars with the % value
  geom_text(
    aes(
      label = scales::percent(premium_pct, accuracy = 1),
      hjust = if_else(premium_pct >= 0, -0.15, 1.15)
    ),
    size = 3, color = "grey20"
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.08, 0.10))
  ) +
  scale_fill_manual(values = c("Positive" = "#2d6a4f", "Negative" = "#c1121f")) +
  labs(
    title    = "AI Wage Premium: Who Benefits Most?",
    subtitle = "% difference in avg. pay between AI and non-AI firms, by industry (2018)",
    x        = "Wage Premium (%)",
    y        = NULL,
    caption  = "Source: 2018 Annual Business Survey (ABS)\nNote: Food Services and Information sectors show the largest wage premiums for AI-adopting firms."
  )


## Employment Scale vs. Wage Premium
ggplot(
  wage_wide |> mutate(premium_sign = if_else(premium_pct >= 0, "Positive", "Negative")),
  aes(
    x = ai_emp_share,
    y = premium_pct,
    size = total_emp,
    color = premium_sign,
    label = naics_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.5) +
  geom_vline(xintercept = 0.15, linetype = "dotted", color = "grey60", linewidth = 0.5) +
  geom_point(alpha = 0.65) +
  geom_text(
    size = 3,
    color = "grey20") +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.05, 0.05)) ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(-0.15, 0.75) ) +
  scale_size_continuous(
    range = c(3, 16),
    labels = scales::label_number(scale = 1e-6, suffix = "M", accuracy = 1),
    name = "Total Employees" ) +
  scale_color_manual(
    values = c("Positive" = "#2d6a4f", "Negative" = "#c1121f"),
    guide = "none") +
  labs(
    title = "Where Does AI Create the Most Value?",
    subtitle = "Share of industry workforce at AI firms vs. wage premium",
    x = "Share of Industry Employees at AI Firms",
    y = "AI Wage Premium",
    caption = "Source: 2018 Annual Business Survey (ABS)") +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    plot.caption = element_text(hjust = 0, color = "grey40", size = 8) )
```

## Takeaways

> **<span style="color: #27485c;"> **AI adoption in 2018 was driven by firm scale, not industry wealth or geography.**

  - Adoption did not follow the money

Management led AI adoption at 7.7% (Figure 1), yet ranked near the bottom in total 2017
revenue (Figure 4). Wholesale Trade and Manufacturing, the two largest industries by revenue, were mid-tier adopters at best. 

::: {.callout-note}
Knowledge-intensive sectors moved first because AI fits their work, not because they had the deepest pockets.
:::

  - Geography is a proxy for industrial composition

California dominates the adoption map, but the random forest assigns **State near-zero permutation importance** once firm characteristics are known. The
geographic pattern reflects where Management, Information, and Professional Services firms
concentrate — not something intrinsic to location.

  - Payroll growth favoured AI-adopting industries — with a COVID caveat

The top four AI adopters grew payroll steadily from 2017 to 2021. Lower-adoption sectors, particularly Agriculture and Arts & Recreation, showed sharp swings around 2019–2020.

  - *The forest confirmed scale as the dominant factor* 

## Machine Learning [(See David's Appendix)](david_appx.qmd)

The single decision tree  splits first on **number of firms ≥ 558**, then on
**employees per firm ≥ 61,** a profile that maps directly onto the concentrated,
knowledge-intensive industries. Adding 2017 panel variables (revenue per firm and average industry wage) cut OOB prediction error from 5.3% to 3.1%, the largest accuracy gain in the analysis.

The Gini vs. permutation comparison shows that Gini inflated the apparent importance of continuous size variables. Permutation importance redistributed credit more evenly and elevated employees-per-firm, suggesting AI adoption reflects a single latent dimension of firm scale rather than any one metric in isolation.

This analysis is observational. Whether AI adoption *caused* the subsequent payroll divergence, or whether the same organizational characteristics that predict adoption also predict payroll growth, cannot be resolved with this data, and remains the natural next step.

