---
title: "David's Appendix "
format: 
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

This is the appendix document for the Random Forest.


```{r setup, echo=FALSE, message=FALSE, warning=FALSE} 
library(here)
library(tidyverse)
library(stringr)
library(scales)
library(ggrepel)
library(tidymodels)
library(ranger)
library(patchwork)
library(forcats)
library(rpart)
library(tibble)   
library(rpart.plot)
library(paletteer)
source(here::here("data", "data_cleaning.R"))
```



## Appendix

### Predicting AI Usage with ML

After doing my EDA in the previous document, I was interested in whether we can predict which firms adopt AI. Thus, I decided to run a Random Forest.

#### Random Forest

The random forest algorithm extends bagging by combining both bagging and feature randomness to create an uncorrelated forest of decision trees. Decision trees work by asking a series of questions that split the data into branches. They are trained using the Classification and Regression Tree (CART) algorithm, with metrics like Gini impurity, information gain, or MSE (mean std. errors) used to evaluate split quality.

#### Building the Forest

While initial results using Gini-based importance suggested that firm size variables overwhelmingly drive AI adoption, this measure is known to be biased toward continuous predictors. Using permutation importance — which evaluates the impact of each variable on model accuracy — provides a more reliable ranking. The results confirm that firm scale remains the primary predictor of AI adoption, followed by workforce intensity and industry sector. In contrast, geographic location (state) contributes no predictive power once firm characteristics are considered.

Because firm size variables are highly correlated, permutation importance likely distributes predictive contribution across them, suggesting that AI adoption is driven by an underlying latent factor of firm scale rather than any single metric.

```{r forest_code}
tidymodels_prefer()

# 2017 industry financials (pre-adoption context)
industry_financials <- cs_panel_clean |>
  filter(year == 2017) |>
  group_by(naics_label) |>
  summarise(
    revenue_per_firm = mean(revenue_per_firm, na.rm = TRUE),
    avg_wage_panel   = mean(avg_wage,         na.rm = TRUE),
    .groups = "drop")

#  9-predictor dataset
rf_data_9 <- tech_2018_clean |>
  filter(
    tech_type == "artificial intelligence",
    used_tech %in% c("Used", "Did not use"),
    !is.na(n_emp), !is.na(payroll), !is.na(n_firms),
    n_emp > 0, payroll > 0, n_firms > 0,
    naics_sector != "Total" ) |>
  mutate(
    ai_adopted   = as.factor(if_else(used_tech == "Used", "Yes", "No")),
    avg_wage     = payroll / n_emp,
    emp_per_firm = n_emp / n_firms,
    naics_label  = as.factor(naics_label),
    state        = as.factor(state) ) |>
  left_join(industry_financials, by = "naics_label") |>
  select(
    ai_adopted,
    # 7-predictor core
    n_emp, n_firms, avg_wage, emp_per_firm, payroll, naics_label, state,
    # +2 from panel
    revenue_per_firm, avg_wage_panel ) |>
  drop_na()

# 7-predictor subset
rf_data_7 <- rf_data_9 |>
  select(-revenue_per_firm, -avg_wage_panel)

cat("Rows in both datasets:", nrow(rf_data_7), "(same rows, different columns)\n")
cat("7-predictor columns:", ncol(rf_data_7) - 1, "\n")
cat("9-predictor columns:", ncol(rf_data_9) - 1, "\n")


#Shared label map
var_labels <- c(
  "n_emp"            = "# Employees",
  "n_firms"          = "# Firms",
  "avg_wage"         = "Avg Wage (2018)",
  "emp_per_firm"     = "Employees / Firm",
  "payroll"          = "Total Payroll",
  "naics_label"      = "Industry Sector",
  "state"            = "State",
  "revenue_per_firm" = "Revenue / Firm (2017)",
  "avg_wage_panel"   = "Avg Industry Wage (2017)")
```

```{r forest_specs}
# settings: 1000 trees, min_n = 2, sqrt(p) predictors tried per split -- AKA random forest
make_spec <- function(importance_type) {
  rand_forest() |>
    set_mode("classification") |>
    set_engine("ranger") |>
    set_args(
      mtry        = NULL,       
      trees       = 1000,
      min_n       = 2,
      probability = FALSE,
      importance  = importance_type)
}

spec_gini  <- make_spec("impurity")    # Gini / mean decrease in impurity
spec_perm  <- make_spec("permutation") # Permutation importance


# Fit all four forests
set.seed(212)
forest_7_gini <- spec_gini |> fit(ai_adopted ~ ., data = rf_data_7)
forest_7_perm <- spec_perm |> fit(ai_adopted ~ ., data = rf_data_7)

set.seed(212)
forest_9_gini <- spec_gini |> fit(ai_adopted ~ ., data = rf_data_9)
forest_9_perm <- spec_perm |> fit(ai_adopted ~ ., data = rf_data_9)


#  OOB error comparison table
oob_summary <- tibble(
  Model      = c("7 predictors", "7 predictors", "9 predictors", "9 predictors"),
  Importance = c("Gini", "Permutation", "Gini", "Permutation"),
  OOB_Error  = c(
    forest_7_gini |> extract_fit_engine() |> pluck("prediction.error"),
    forest_7_perm |> extract_fit_engine() |> pluck("prediction.error"),
    forest_9_gini |> extract_fit_engine() |> pluck("prediction.error"),
    forest_9_perm |> extract_fit_engine() |> pluck("prediction.error"))) |>
  mutate(OOB_Accuracy = 1 - OOB_Error)

print(oob_summary)
``` 

```{r forest_vizs}
# OOB error
oob_summary |>
  ggplot(aes(x = Model, y = OOB_Error, fill = Importance)) +
  geom_col(position = "dodge", width = 0.55) +
  geom_text(
    aes(label = scales::percent(OOB_Error, accuracy = 0.1)),
    position = position_dodge(width = 0.55),
    vjust = -0.5, size = 3.5) +
  scale_fill_viridis_d(option = "E", begin = 0.2, end = 0.8) +   
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.1))) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "OOB Prediction Error: 7 vs 9 Predictors",
    subtitle = "Lower is better — error rate on out-of-bag samples",
    x        = NULL, y = "OOB Error Rate", fill = "Importance method",
    caption  = "Source: 2018 Annual Business Survey + 2017 Economic Census")+
   theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "right")


# Importance extraction helper
get_importance <- function(forest, method_label, n_pred_label) {
  forest |>
    extract_fit_engine() |>
    pluck("variable.importance") |>
    enframe(name = "Variable", value = "Importance") |>
    mutate(
      Variable   = recode(Variable, !!!var_labels),
      Method     = method_label,
      Predictors = n_pred_label,
      # Normalise within each forest so Gini and Permutation scales are comparable
      Importance = Importance / max(abs(Importance)))
}

importance_all <- bind_rows(
  get_importance(forest_7_gini, "Gini",        "7 predictors"),
  get_importance(forest_7_perm, "Permutation",  "7 predictors"),
  get_importance(forest_9_gini, "Gini",        "9 predictors"),
  get_importance(forest_9_perm, "Permutation",  "9 predictors"))

# Importance comparison helper
plot_importance_comparison <- function(data, title_suffix) {
  data |>
    select(Variable, Method, Importance) |>
    pivot_wider(names_from = Method, values_from = Importance) |>
    mutate(Variable = fct_reorder(Variable, Gini)) |>
    ggplot(aes(y = Variable)) +
    geom_segment(
      aes(x = Permutation, xend = Gini, yend = Variable),
      color = "grey70", linewidth = 0.6) +
    geom_point(aes(x = Gini,        color = "Gini"),        size = 3) +
    geom_point(aes(x = Permutation, color = "Permutation"), size = 3) +
    scale_color_viridis_d(option = "E", begin = 0.2, end = 0.8) + # ← changed
    theme_minimal(base_size = 11) +
    labs(
      title    = glue::glue("Gini vs Permutation Importance — {title_suffix}"),
      subtitle = "Normalised 0–1 within each method. Dots far apart = Gini bias.",
      x = "Normalised importance", y = NULL, color = NULL,
      caption = "Source: 2018 Annual Business Survey") +
    theme(legend.position = "bottom")
}

#  importance plot
importance_all |>
  select(Variable, Method, Importance, Predictors) |>
  pivot_wider(names_from = Method, values_from = Importance) |>
  mutate(
    x_start  = ifelse(Gini > Permutation, Gini - 0.02, Gini + 0.02),
    x_end    = ifelse(Gini > Permutation, Permutation + 0.02, Permutation - 0.02),
    Variable = paste0(Variable, "___", Predictors),
    Variable = fct_reorder(Variable, Permutation)) |>
  ggplot(aes(y = Variable)) +
  geom_segment(
    aes(x = x_start, xend = x_end, yend = Variable),
    arrow = arrow(length = unit(0.18, "cm"), type = "closed", ends = "last"),
    color = "grey60", linewidth = 0.6) +
  geom_point(aes(x = Gini,        color = "Gini"),        size = 2.5) +
  geom_point(aes(x = Permutation, color = "Permutation"), size = 2.5) +
  scale_color_viridis_d(option = "E", begin = 0.2, end = 0.8) +
  scale_y_discrete(labels = ~ sub("___.*", "", .x)) +      # ← strip the key suffix
  facet_wrap(~ Predictors, scales = "free_y", ncol = 1) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text      = element_text(face = "bold"),
    plot.title      = element_text(face = "bold", size = 16),
    plot.subtitle   = element_text(color = "grey30"),
    legend.position = "right") +
  labs(title    = "Variable Importance: All Four Models",
    subtitle = "Arrow points Gini → Permutation; longer arrow = greater Gini overestimation",
    x = "Normalised importance", y = NULL, color = NULL,
    caption  = "Source: 2018 Annual Business Survey + 2017 Economic Census")+
   theme(plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "grey30"),
    legend.position = "right")
  


# Single decision tree 
# The forest uses 1000 of these, each on a bootstrap sample with random feature
# subsets — individual trees are weak but their average is robust.

single_tree <- decision_tree() |>
  set_mode("classification") |>
  set_engine("rpart") |>
  set_args(cost_complexity = 0.02, tree_depth = 4) |>
  fit(
    ai_adopted ~ n_firms + emp_per_firm + avg_wage + payroll + n_emp,
    data = rf_data_7)   # no state & naics_label to keep the tree readable
  

single_tree |>
  extract_fit_engine() |>
  rpart.plot(
    type        = 4,
    extra       = 104,          # show % and n at each leaf
    box.palette = "PuBu",
    roundint    = FALSE,
    fallen.leaves = TRUE,      
    main = paste(
      "One example tree (depth 4, cost_complexity = 0.02)",
      "The random forest grows 1000 of these on bootstrapped samples",
      "Each split considers only sqrt(5) ≈ 2 randomly chosen predictors",
      sep = "\n"))


# OOB Confusion matrices
# (using the permutation forests since they have more honest importance estimates)
cat("\n── Confusion matrix: 7-predictor forest (permutation) ──\n")
forest_7_perm |>
  extract_fit_engine() |>
  pluck("confusion.matrix") |>
  t() |>
  print()

cat("\n── Confusion matrix: 9-predictor forest (permutation) ──\n")
forest_9_perm |>
  extract_fit_engine() |>
  pluck("confusion.matrix") |>
  t() |>
  print()
```


